---
title: "3. A genome-resolved community"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. A genome-resolved community}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Vignettes 1 and 2 worked on one genome. This one asks the questions that only
make sense across several: how redundantly is a capability provided, which
genome is its only provider, and which chains does the community complete that
no single member does.

```{r}
library(gifter)
```

## The community

Four illustrative genomes, built around a real curated chain: the stepwise
breakdown of arabinoxylan, a plant cell-wall polysaccharide that reaches the
large intestine undigested.

```{r}
markers <- function(namespace, accession) {
  data.frame(
    gene_id = paste0("gene_", seq_along(accession)),
    namespace = namespace, accession = accession,
    stringsAsFactors = FALSE
  )
}

# A: strips arabinose side chains off the xylan backbone.
debrancher <- evaluate_gifts(markers("KO", "K01209"))

# B: cleaves the exposed backbone and releases xylose outside the cell.
backbone_degrader <- evaluate_gifts(markers("CAZY", c("GH11", "GH39")))

# C and D: take up free xylose and catabolise it. Two genomes, same capability.
consumer <- evaluate_gifts(markers(
  "KO", c("K10543", "K10544", "K10545", "K01805", "K00854")
))
```

Each supports exactly what its name says:

```{r}
lapply(
  list(A = debrancher, B = backbone_degrader, C = consumer),
  function(x) x$gifts$gift_id[x$gifts$complete]
)
```

These four capabilities are connected in the curated database already — this is
not an arrangement invented for the tutorial:

```{r}
chain <- c("arabinoxylan_debranching", "xylan_degradation",
           "xylose_uptake_abc", "xylose_degradation_isomerase")
edges <- gift_graph()
edges[edges$from_gift %in% chain & edges$to_gift %in% chain,
      c("from_gift", "shared_anchor", "to_gift")]
```

## Step 1: bind the genomes together

```{r}
community <- gifter_community(
  A = debrancher,
  B = backbone_degrader,
  C = consumer,
  D = consumer
)
community
```

The argument names become the genome identifiers, and the container refuses two
things that would silently produce nonsense: genomes evaluated against different
database releases, and an abundance vector that does not name exactly these
genomes.

## Step 2: how is capability distributed?

```{r}
traits <- community_traits(
  community,
  universes = list(gift_universe(label = "all curated GIFTs"))
)
metrics <- traits$metrics
```

Start with the community as a whole:

```{r}
metrics[metrics$target_type == "community",
        c("metric_id", "value", "numerator", "denominator")]
```

Four capabilities are present somewhere, while the average genome carries 1.5.
Both numbers are reported rather than only their ratio, because the ratio alone
hides which of the two moved.

Now per capability — how many genomes provide each?

```{r}
providers <- metrics[metrics$metric_id == "provider_count", ]
providers[, c("target_id", "value", "denominator")]
```

Two of the four capabilities rest on a single genome. That is the
`singleton_fraction` above: 0.5. Which genomes are those single points?

```{r}
unique_rows <- metrics[metrics$metric_id == "unique_contribution", ]
unique_rows[, c("target_id", "value")]
```

A and B each provide something nobody else does. C and D are duplicates of each
other, so neither is anyone's only provider.

Read that last sentence carefully. `unique_contribution` is a property of **this
sample**, not of the organism. Sequence one more xylose consumer and C and D
stay at zero; miss B in your assembly and its unique contribution disappears
along with it.

## Step 3: presence and abundance are different questions

If you have relative abundances, supply them — but note that gifter keeps them in
separate rows from the presence-based counts, deliberately.

```{r}
weighted <- gifter_community(
  A = debrancher, B = backbone_degrader, C = consumer, D = consumer,
  abundance = c(A = 0.05, B = 0.10, C = 0.35, D = 0.50)
)
weighted_metrics <- community_traits(
  weighted, universes = list(gift_universe(label = "all curated GIFTs"))
)$metrics

comparison <- weighted_metrics[
  weighted_metrics$metric_id %in% c("provider_count", "abundance_coverage"), ]
comparison[, c("target_id", "metric_id", "value")]
```

Xylan degradation has one provider carrying 10% of the community; xylose uptake
has two carrying 85%. Those are different facts and neither substitutes for the
other. A capability held only by a rare genome is a different situation from one
held only by a dominant genome, and collapsing them into a single "importance"
score would destroy the distinction.

`abundance_coverage` is the fraction of observed genome abundance carrying the
capability. It is **not** a fraction of activity, transcription, flux or
ecological effect.

## Step 4: who could hand what to whom

```{r}
network <- community_network(community)
network
```

```{r}
network$edges
```

Three edges: A feeds B, and B feeds both consumers. Each edge names the two
GIFTs and the anchor molecule that connects them, so it can be checked.

```{r}
network$nodes
```

B is the hub — it receives from A and provides to both C and D.

### The rule that stops this from being nonsense

C and D both encode xylose uptake *and* xylose catabolism, and those two are
connected in the GIFT graph through `XYLOSE_IN`. So why is there no C → D edge?

```{r}
anchors <- get_gift_anchors("xylose_uptake_abc")
anchors[, c("anchor_id", "role", "compartment")]
```

`XYLOSE_IN` is **cytoplasmic**. A molecule inside a cell cannot be handed to
another organism, so a cross-genome edge requires the producing GIFT's output
anchor to be declared `extracellular`. Without that rule the network would
assert that one bacterium passes another a metabolite that never crosses a
membrane.

The same link inside a single genome is an ordinary composition step, and it is
still reported as one:

```{r}
coverage <- network$chain_coverage
coverage[, c("from_gift", "shared_anchor", "status", "within_genome")]
```

Four statuses are possible; three of them appear here.

- `within_genome` — some genome holds both ends. C and D each complete the
  uptake-to-catabolism step internally.
- `community_distributed` — no genome holds both ends, both are present, and the
  molecule is extracellular. **This is the interesting one**: the community
  completes a chain no member completes.
- `not_represented` — one half is missing entirely. A releases arabinose and
  nobody in this community takes it up.
- `not_transferable` — the two halves are in different genomes but the molecule
  stays inside a cell, so nothing completes the link. It does not arise above,
  because C and D each hold both halves. Split them and it does:

```{r}
uptake_only <- evaluate_gifts(markers("KO", c("K10543", "K10544", "K10545")))
catabolism_only <- evaluate_gifts(markers("KO", c("K01805", "K00854")))

split_coverage <- community_network(
  gifter_community(U = uptake_only, V = catabolism_only)
)$chain_coverage

split_coverage[split_coverage$shared_anchor == "XYLOSE_IN",
               c("status", "providers", "recipients")]
```

  Both halves are encoded, by different organisms, and the community still
  completes nothing: cytoplasmic xylose does not travel. Reporting that as
  `community_distributed` would be the error the status exists to prevent.

### What an edge is not

An edge is a **potential** compatibility relationship between what one genome
encodes and what another encodes. It is not evidence that:

- either GIFT is expressed;
- the molecule is released at a useful concentration;
- the two organisms co-occur in space or time;
- exchange actually happens.

Cross-feeding is a hypothesis an edge can support. It is never a conclusion an
edge establishes. gifter will not name it cross-feeding, and neither should a
figure legend built on this output.

## Step 5: fragmented genomes in a community

Real communities are assembled from MAGs of uneven quality. As in vignette 2,
completeness changes how a *negative* call is read — and here it is resolved per
genome, so one fragmented member does not contaminate everyone's denominators.

```{r}
mixed <- gifter_community(
  A = debrancher, B = backbone_degrader, C = consumer,
  quality = c(A = 0.98, B = 0.96, C = 0.45),
  policy = "completeness", threshold = 0.9
)

mixed_metrics <- community_traits(
  mixed, universes = list(gift_universe(label = "all curated GIFTs"))
)$metrics

fractions <- mixed_metrics[mixed_metrics$metric_id == "provider_fraction", ]
fractions[, c("target_id", "value", "numerator", "denominator")]
```

`arabinoxylan_debranching` is divided by 2, not 3: C is too fragmented to say it
lacks debranching, so it is not counted as a genome that was asked. Meanwhile C's
own positive calls survive untouched — quality only ever changes the reading of
absence.

```{r}
sum(mixed$matrix[, "C"] %in% TRUE)
```

## Putting the three tutorials together

```text
annotation table
      |  evaluate_gifts()
      v
  GIFT calls  ------------------------------> trace_gift()
      |
      |  genome_traits()          within a declared gift_universe()
      v
genome traits
      |
      |  gifter_community() + community_traits()
      v
community traits  ---------------------------> community_network()
```

At every step the same discipline applies: the number carries the set it was
counted over, the evidence that produced it stays attached, and nothing becomes
more certain than the calls underneath. A GIFT call is what a genome encodes —
not what an organism is doing.
