---
title: "How a GIFT is built"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How a GIFT is built}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A GIFT is not imported from a pathway database or assembled from a convenient
gene list. It is a biological capability with a stated boundary and an explicit
completeness model. Building one is therefore a curation process first and a
database-compilation process second.

```text
candidate capability
  -> prior-attempt search
  -> precise claim and GIFT type
  -> boundaries or machinery architecture
  -> routes/reactions or functions/systems        (reference knowledge bases)
  -> jointly required components
  -> sufficiently specific marker evidence        (catalogue-wide specificity screens)
  -> test against genomes                         (KEGG reference genomes + named controls)
  -> composition and redundancy review
  -> provenance, validation and tests
  -> implemented, deferred or refused outcome
```

This article explains each stage, which data it draws on, and where the record
of every decision can be found. The normative contract lives in the
[development and architecture guide](https://github.com/alberdilab/gifter/blob/main/inst/doc/architecture.md#curating-a-new-or-changed-gift).
Technical terms are defined in the [glossary](glossary.html).

## Which data are used, and are they always the same?

Curation draws on four kinds of data. They do different jobs, and only some of
them are the same from one GIFT to the next.

| Data | Same for every GIFT? | Used to | Never used to |
|---|---|---|---|
| **Reference knowledge bases**: Rhea, ChEBI, KEGG, MetaCyc, UniProt, NCBIfam, dbCAN, literature | Yes, at a pinned release | define the chemistry, the molecules, the enzymes and the candidate markers | decide where a GIFT's boundary lies; that is gifter's own decision |
| **Catalogue-wide specificity screens** | Yes; rerun when a resource is updated | measure how many distinct reactions each marker can license | admit a marker on their own; they flag rows for a curator to read |
| **KEGG reference genome set**: every genome in the KEGG catalogue, about 11,900 | Yes; the snapshot date is recorded with each assessment | count how many genomes complete a candidate, which function near-misses lack, and whether it is distinguishable from homologous machinery | define a GIFT by its prevalence |
| **Named control genomes** | **No.** Chosen for each investigation | check the definition against organisms whose biology is established in the literature | produce a call; the expectation is compared with the marker result, never fed into it |

Two further kinds of data meet gifter only after curation and never define a
GIFT: observed phenotypes (BacDive, MediaDive and species-level trait tables)
and application datasets such as metagenome-assembled genomes. They are
described in [Data that do not define GIFTs](#data-that-do-not-define-gifts).

The releases the current database was compiled against are recorded in the
database itself:

```{r}
library(gifter)
release <- gifter_db_version()
data.frame(field = names(release), value = unlist(release, use.names = FALSE))
```

## Begin with the question, not the markers

The first curation statement is one sentence describing what a positive call
means. It should name an encoded capability that is useful to compare among
genomes without claiming expression, activity, physiological state or
phenotype.

Before new evidence is collected, the curator searches the append-only
[catalogue-expansion attempt log](https://github.com/alberdilab/gifter/blob/main/inst/doc/catalogue-expansion-attempts.tsv)
for the candidate, its aliases and the wider request that produced it, and reads
the [deferral register](https://github.com/alberdilab/gifter/blob/main/inst/doc/deferral-register.md)
for standing blockers. A previous refusal is reopened only when its evidence or
recorded retrigger has changed. A new bounded investigation is then registered
as `in_progress`, even if it may ultimately add nothing. This makes unsuccessful
searches and honest refusals as discoverable as additions.

## Choose the completeness model

Every GIFT declares one `gift_type`. The type decides what must be complete and
which source tables may describe the capability.

| Type | A positive call requires at least one complete... | Principal boundary |
|---|---|---|
| `metabolic` | route | declared input and output anchors |
| `structural` | architecture | machinery needed to build a defined structure |
| `regulatory` | circuit | machinery needed to sense a signal and execute a response |
| `defense` | mechanism | machinery needed to execute a defined defense mechanism |

A type is not a display category. A structure is not forced into the metabolic
model by inventing a boundary molecule, and a metabolic transformation is not
represented as a generic machinery checklist.

## Define the claim and its boundaries

For a metabolic GIFT, the input and output anchors identify where the capability
begins and ends. Good boundaries are meaningful environmental or host
compounds, precursors, stable products or branchpoints. Internal reaction
participants are not automatically anchors, and only shared declared anchors
connect GIFTs.

For structural, regulatory and defense GIFTs, the analogous decision is the
scope of the machine: which alternative complete implementations exist and
which functions every implementation requires. The claim should be narrow
enough that the genomic evidence can support it, but large enough to represent
a biologically meaningful capability rather than an arbitrary protein or
single reaction.

At this stage the curator also checks whether the proposal duplicates an
existing GIFT, bundles independently meaningful capabilities, or is already
represented by a chain of composable metabolic GIFTs.

## Where the biological knowledge comes from

Each external resource is trusted for one thing. The table states what each
contributes and what it is not allowed to decide.

| Resource | Supplies | Is not trusted for |
|---|---|---|
| [Rhea](https://www.rhea-db.org/) | reaction identity: every reaction is a Rhea master reaction where Rhea covers the chemistry, with its participants | the direction a route uses it, which gifter curates per route |
| [ChEBI](https://www.ebi.ac.uk/chebi/) | the identity of each anchor molecule | whether that molecule is a meaningful boundary |
| [KEGG](https://www.kegg.jp/) modules and orthology | a starting pathway skeleton and KO markers | GIFT boundaries: modules are read and cut, never adopted whole |
| [MetaCyc](https://metacyc.org/) | reconnaissance of pathway variants and alternative routes | evidence from pathway pages that could not be read without a subscription |
| [UniProt](https://www.uniprot.org/) and primary literature | characterised enzymes and subunits, and which proteins a complex jointly requires | a family-wide claim from one characterised member |
| [NCBIfam](https://www.ncbi.nlm.nih.gov/genome/annotation_prok/evidence/) (`hmm_PGAP` 20.0) | HMM profile markers; only the `equivalog` and `equivalog_domain` grades are admitted | anything a `subfamily` or `domain` profile would imply |
| [dbCAN](https://bcb.unl.edu/dbCAN2/) (`db_v5-2-9_5-5-2026`) | CAZy family, subfamily and eCAMI-cluster markers, with their substrate and EC associations | a substrate-specific claim from a polyspecific family |

KEGG modules are the most common starting point and the most common place where
gifter departs from its source. For the guanylate GIFT, for example, KEGG
module M00050 runs from IMP all the way to GTP. gifter ends the GIFT at GMP,
the first product that establishes guanylate identity, and leaves out the
shared kinase steps after it. Every such departure is written down, either in
[`SOURCES.md`](https://github.com/alberdilab/gifter/blob/main/inst/extdata/database-source/SOURCES.md)
or, when it changes which markers count, in the database's own change history.

## Materialize every valid implementation

The Boolean layers are deliberately explicit. For a metabolic GIFT:

```text
GIFT = ANY complete route
route = ALL required reactions
reaction = ANY complete enzyme system
system = ALL required components
component = ANY accepted marker
```

Routes are minimal complete biochemical paths between the declared anchors.
Each reaction is identified by its Rhea master where possible, while the
reaction's `forward` or `reverse` use in the route is recorded separately.
Alternative routes, non-homologous enzyme replacements, monomeric and
multisubunit enzymes, and protein fusions are represented in their proper
layers instead of being compressed into a Boolean expression.

Structural, regulatory and defense GIFTs use the same Boolean shape under
biologically appropriate names:

```text
GIFT = ANY complete architecture, circuit or mechanism
implementation = ALL required functions
function = ANY complete system
system = ALL required components
component = ANY accepted marker
```

Accessory functions may be recorded, but they do not silently become required.
For an incomplete call, this structure is what lets gifter report the closest
implementation and the missing reactions or functions.

## Choose marker evidence

Markers support components; they never support a reaction, route or GIFT
directly. Marker identity is the pair `namespace + accession`, such as a KEGG
Orthology accession or a versioned NCBIfam profile.

The governing test is specificity: the biological claim must not be more
specific than the marker evidence. A broad enzyme family cannot establish a
substrate-specific reaction, a generic chemoreceptor cannot identify an
effector, and a profile shared by homologous machines cannot distinguish those
machines. The rest of a route may provide context for a genuinely broad shared
reaction, but it cannot repair a marker that does not identify its own curated
component.

### Specificity screens across the whole catalogue

Specificity is not judged one accession at a time from its name. Three screens
measure it across each marker namespace. They use the reference knowledge bases
only, not genomes:

| Namespace | What is measured | Script |
|---|---|---|
| KEGG orthology | each KO's EC numbers are followed to the Rhea master reactions they reach. A KO reaching one master reaction states one reaction; a KO reaching several states only their union | [`marker_specificity_screen.R`](https://github.com/alberdilab/gifter/blob/main/data-raw/marker_specificity_screen.R) |
| NCBIfam | the same join for every equivalog profile, after the grade filter. It also asks whether a profile resolves a curated reaction that no single-reaction KO resolves; 229 profiles reaching a curated reaction are refused on grade alone | [`ncbifam_equivalog_screen.R`](https://github.com/alberdilab/gifter/blob/main/data-raw/ncbifam_equivalog_screen.R) |
| CAZy (eCAMI clusters) | the fraction of a cluster's EC-annotated members that carry the claimed EC | [`extract_cazy_subfamilies.R`](https://github.com/alberdilab/gifter/blob/main/data-raw/extract_cazy_subfamilies.R) |

A screen finds the rows that need a curator; it does not decide them. A KO
reaching several reactions may group genes of different activities, which
cannot support a substrate-specific claim. Or it may be one genuinely
promiscuous enzyme, which supports every reaction it performs. Reading which
participants vary between the reactions usually settles it.

The screen outputs are kept in
[`data-raw/reference/`](https://github.com/alberdilab/gifter/tree/main/data-raw/reference),
whose README explains every table.

### Confidence grades

Every accepted marker row records an evidence type, a confidence grade, the
source it was taken from, and any ambiguity that affects interpretation.

| Grade | Meaning |
|---|---|
| `curated` | the marker identifies the component. Most KO rows; for a CAZy cluster, at least 93% EC agreement across ten or more annotated members |
| `high-confidence` | strong but not exhaustive support. Admitted NCBIfam equivalogs; CAZy clusters with at least 70% EC agreement |
| `putative` | accepted with a stated doubt, recorded in the row's notes. The archaellum ATPase KO is one: it also collects the ATPases of homologous archaeal pili |
| `ambiguous` | the only available evidence, and polyspecific; for a CAZy cluster, 50--70% EC agreement |

Below 50% EC agreement a CAZy cluster is not admitted at all: a cluster in which
3 of 96 annotated members carry the claimed EC is evidence against the
assignment, not weak evidence for it.

Confidence reaches the call. `evaluate_gifts()` reports `evidence_confidence`,
the weakest grade among the markers supporting the best route, so a GIFT resting
on an ambiguous family never reads like one resting on curated orthology. A
marker that cannot support the claim is refused rather than stretched to fit it.

NCBIfam accessions keep their version suffix (`NF040708.3`), because a rebuilt
profile may match different sequences. The same caution applies to dbCAN's
eCAMI cluster identifiers, which are renumbered between releases. Annotate
genomes against the releases the database names, or CAZy markers will fail to
match without any error.

## Test the definition against genomes

A definition that reads well on paper still has to behave sensibly on real
genomes. Two genome datasets are used for this, and they answer different
questions.

### The KEGG reference genome set: the same for every GIFT

The population-level reference is the KEGG genome catalogue: about 11,900
genomes, of which roughly 10,100 are bacterial and 470 archaeal. For each
marker, KEGG's REST service returns every genome carrying it, so a candidate's
marker set can be evaluated across the whole catalogue without annotating a
single genome. The same reference serves every curation round. The snapshot
date and the exact denominators are stated in each assessment, because KEGG
grows between snapshots. Domain counts come from KEGG's organism hierarchy,
because the catalogue's own prokaryote flag admits some eukaryotes.

The screen answers four questions:

1. **How many genomes complete the candidate?** A GIFT that no genome completes
   is refused on evidence. Vitamin C biosynthesis was refused this way: 0 of
   10,151 bacterial genomes completed it.
2. **What do near-misses lack?** If genomes one function short mostly miss the
   same component, that component may be an annotation gap or a taxon-specific
   replacement still to be curated, rather than real absence.
3. **Is the candidate distinguishable from homologous machinery?** Secretion
   systems share ancestry with the flagellum and with pili, so the screen counts
   how many genomes complete the homologue while supporting none of the
   candidate's roles.
4. **Does each required component do discriminating work?** A required
   component that almost every candidate genome carries anyway adds no
   specificity.

Prevalence never decides on its own. A rare architecture can be admitted when
its markers resolve it: the *Bacillus subtilis* W23-type ribitol-phosphate wall
teichoic acid architecture completes in only 22 genomes and was curated under an
explicit biological-importance exception.

This reference has two limits, and both are stated wherever it is used:

- **It is KO-only.** CAZy, NCBIfam and Pfam markers cannot be reached through
  it, so a GIFT whose usual evidence is a CAZy family is under-counted here.
- **It is KEGG's own per-genome assignment**, not the output of the annotation
  pipeline a user would run. Its counts are therefore an upper bound on what a
  user's annotation will support.

KEGG's per-genome assignments may not be redistributed, so only aggregate
counts are committed. The per-genome data stay in a local cache that each
script rebuilds.

When a KO is suspected of merging homologues, the screen goes down to sequence
level. It retrieves the exact proteins KEGG assigned to that KO and searches
them with the candidate NCBIfam equivalogs, plus contrast profiles for the
homologous machines, at each profile's own gathering threshold. Contrast
profiles only name what a protein is; they are never accepted as gifter
evidence. The archaellum, wall teichoic acid and type II secretion GIFTs were
all checked this way.

### Named control genomes: chosen for each investigation

The second dataset is small and changes every time. For each candidate the
curator picks organisms whose relevant biology is established in the
literature, both organisms expected to carry the capability and organisms
expected to lack it. Examples include:

- 16 gut and model organisms for short-chain fatty acid formation;
- 18 genomes for vitamin biosynthesis;
- three archaellated archaea and two flagellated bacteria for the archaellum;
- 12 strains for type II secretion: seven expected to carry it, three expected
  to lack it, and two marked `unverified`;
- 17 strains for type III and type VI secretion.

The controls are listed by KEGG organism code in the assessment, or in a
`*-controls.tsv` table beside the screen.

A control's expected result is written beside the marker result and compared
with it. It is never used to produce the call, and a strain whose expectation
was not actually checked is marked `unverified`. A control that disagrees is
reported as it is, not fixed: it may point to a missing route (see
[acetate](#a-metabolic-gift-acetate-interconversion) below), a marker gap, or a
genuine exception.

## Data that do not define GIFTs

Some data are used to evaluate gifter but never to define a GIFT.

- **Observed phenotypes.** BacDive substrate-use records, MediaDive growth media
  and species-level trait tables are compared with gifter's calls as a
  benchmark. They are not accepted as marker evidence: an organism growing on a
  substrate says nothing about which gene does the work.
- **Application datasets.** The GTDB panel, the chicken caecal
  metagenome-assembled genomes and the Microflora Danica samples used in the
  manuscript analyses are readings of the catalogue, not inputs to it.

A benchmark can, however, raise a question for curation. When genomes of
organisms observed to perform a capability are called unsupported, gifter's own
trace names the reaction that failed. That lead then goes through the full
process above.

- **GlcNAc degradation.** Of 119 strains observed to use
  N-acetylglucosamine as a carbon source, only 15% were called positive. The
  same kinase reaction failed in all 30 sampled failures. The boundaries were
  right but a whole route was missing, and a second route through the
  phosphotransferase system was curated in database 2026.26.1.
- **Serine biosynthesis.** The lead's premise did not survive measurement, so
  nothing was changed.

A benchmark lead is a question, not a diagnosis. The record is in
[the phenotype curation leads](https://github.com/alberdilab/gifter/blob/main/inst/doc/proposal-phenotype-curation-leads.md).

## Two worked examples

### A metabolic GIFT: acetate interconversion

The candidate was "acetate formation", assessed with the short-chain fatty
acids on 2026-08-18.

- **Knowledge bases.** KEGG module M00579 offered a clean two-reaction
  skeleton: phosphate acetyltransferase then acetate kinase, both with Rhea
  master reactions.
- **Specificity.** The KOs for both enzymes each state one reaction.
- **Reference genomes.** 6,582 KEGG genomes carry both enzymes.
- **Controls.** Every reference genome checked was positive except
  *Akkermansia muciniphila*, which excretes acetate but lacks acetate kinase.
  That was recorded as a missing alternative route, not a reason to refuse.
- **The claim.** Both reactions are reversible and the markers cannot tell the
  direction, so the GIFT was named an *interconversion* rather than a
  *formation*. A name that claims direction would exceed the evidence.
- **A refusal in the same assessment.** The butyrate-kinase route was refused.
  Its enzymes are shared with *Bacillus* branched-chain metabolism at both the
  KO and TIGRFAM levels, so no change of marker namespace could make it
  specific.

The resulting definition can be read with the public accessors:

```{r}
reactions <- get_gift_reactions("acetate_interconversion")
reactions[, c("route_id", "reaction_id", "orientation")]

systems <- get_reaction_systems("RHEA:19521")
systems[, c("component_id", "namespace", "accession", "confidence", "source")]

evidence <- get_gift_evidence("acetate_interconversion")
evidence[, c("evidence_kind", "location")]
```

The `source` column names the module and date the markers were taken from. The
evidence row points to the assessment that defines and defends the GIFT.

### A structural GIFT: the type II secretion system

The type II secretion system was deferred by the 2026-10-03 survey of
structural machines. Its prepilin peptidase is not merely a homologue of the
type IVa pilus peptidase; it is often the same enzyme. Once the rule for shared
components was settled, the curation questions were quantitative, and a
dedicated screen answered them over 11,908 KEGG genomes:

- **Is it a distinct machine?** 1,571 genomes complete the core Gsp
  apparatus. 703 complete it without the type IVa pilus, and 674 complete the
  pilus without it, so the apparatus is not a proxy for the pilus.
- **Does the shared peptidase discriminate?** Requiring it removes only 35 of
  the 1,571 genomes, so it does almost no discriminating work. It could not be
  refused either: 1,319 of the 1,536 complete apparatuses reach the peptidase
  through PilD rather than GspO.
- **Controls.** Twelve named strains were evaluated.
  - Five of the seven expected to carry a type II system complete it,
    including *Pseudomonas aeruginosa* PAO1 and *Escherichia coli* K-12.
  - *Vibrio cholerae* N16961 and *Legionella pneumophila* fall one role short
    (GspD and GspC respectively). Both shortfalls are kept in the table as
    observations about the markers.
  - The three expected to lack it (*Bacillus subtilis* 168, *Campylobacter
    jejuni* and *Helicobacter pylori*) support at most two of the eleven roles.

Each Gsp role is supported by a KEGG orthology and, where one exists, an NCBIfam
equivalog for the same role:

```{r}
machinery <- get_gift_machinery("type_ii_secretion_system")
unique(machinery[, c("function_name", "required")])
head(machinery[, c("component_name", "namespace", "accession", "confidence")])

get_gift_evidence("type_ii_secretion_system")[, c("evidence_kind", "location")]
```

Here the evidence lists the analysis script and the three result tables it
wrote, as well as the curation documents. The numbers above can be regenerated
with `Rscript data-raw/t2ss_prevalence.R`, or checked from the cached inputs
with `--offline`.

## Check composition and traceability

The proposed GIFT is inspected against the existing catalogue. Metabolic
composition may occur only through declared anchors; a shared internal
metabolite does not create an edge. Machinery GIFTs remain independent even
when they genuinely share a component.

The complete evidence path must remain recoverable:

```text
observed marker
  -> component
  -> system
  -> reaction or function
  -> route, architecture, circuit or mechanism
  -> GIFT
```

When a caller supplies gene identifiers, the trace continues to those genes.
If any level cannot be explained, the definition is not ready to compile.

## Record provenance, build and close the attempt

The reviewable source of truth is the collection of TSV files under
[`inst/extdata/database-source`](https://github.com/alberdilab/gifter/tree/main/inst/extdata/database-source).
SQLite is generated from those files and is never edited by hand.

For a metabolic GIFT, source records are normally added in dependency order:

1. GIFT metadata and any new anchors;
2. anchor roles and related external pathway links;
3. reactions and reaction cross-references;
4. routes and ordered route reactions;
5. enzyme systems and required components;
6. reusable markers and component-to-marker evidence;
7. curation evidence: the proposal that defines the GIFT and any analysis
   script and result table behind it, in `gift_evidence.tsv`;
8. provenance, biological change history and release metadata.

The machinery models use the corresponding architecture, circuit or mechanism
tables and their typed function, system, component and marker tables.

Validation checks identifiers, foreign keys, type-specific rules, Boolean-layer
completeness, marker admission rules and release metadata before SQLite is
compiled:

```sh
Rscript data-raw/build_database.R
Rscript -e 'testthat::test_local(".")'
```

Tests then exercise biological behaviour: alternatives under OR, jointly
required elements under AND, incomplete calls, specificity failures,
composition, and evidence tracing. The resulting definition is inspected with
the public accessors and in the [Atlas](../atlas/).

Finally, the investigation is closed with one of three explicit outcomes:

- **implemented:** exact new or changed `gift_id` values and their database
  change record;
- **deferred:** a standing blocker and the evidence that would retrigger work;
- **refused:** the claim cannot be supported without exceeding the evidence or
  violating the ontology.

A refusal is a curation result, not missing data. Recording it prevents a later
curator from unknowingly repeating the same investigation or admitting the same
over-broad evidence under a different name.

### Where to find the record of any decision

| Question | Where it is answered |
|---|---|
| Which documents, scripts and result tables define this GIFT? | `get_gift_evidence()`, and the "Curation evidence" section of the GIFT's page in the [Atlas](../atlas/#gifts) |
| Why is the GIFT bounded where it is, and what was refused alongside it? | the curation proposal or assessment the evidence points to, under [`inst/doc/`](https://github.com/alberdilab/gifter/tree/main/inst/doc) |
| What was investigated, when, and with what outcome? | the [catalogue-expansion attempt log](https://github.com/alberdilab/gifter/blob/main/inst/doc/catalogue-expansion-attempts.tsv) |
| Why is a candidate not in the catalogue, and what would change that? | the [deferral register](https://github.com/alberdilab/gifter/blob/main/inst/doc/deferral-register.md) |
| Which exact genome screens and controls were used? | the analysis scripts in [`data-raw/`](https://github.com/alberdilab/gifter/tree/main/data-raw) and their outputs in `data-raw/reference/` |
| What changed in the biological content, and which GIFTs did it affect? | `database_changelog()`, and the [Changes view of the Atlas](../atlas/#changelog) |
| Which upstream releases and pathway modules were used? | `gifter_db_version()` and [`SOURCES.md`](https://github.com/alberdilab/gifter/blob/main/inst/extdata/database-source/SOURCES.md) |

## What database compilation does—and does not do

Compilation validates and transforms reviewed source tables into the packaged
SQLite database. It does not discover biological boundaries, infer missing
routes, upgrade uncertain markers or decide that a candidate is a GIFT. Those
are curation decisions and must already be explicit, sourced and testable before
the compiler runs.
