---
title: "1. Evaluating a genome"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Evaluating a genome}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This is the first of three tutorials and covers the whole of one analysis: from
an annotation table to a set of calls, and from a call back down to the genes
that produced it.

```{r}
library(gifter)
```

## What gifter answers, and what it does not

A **GIFT** is a genome-inferred functional trait: a biologically meaningful
capability whose genomic support is decided by an explicit, curated model.
gifter answers one question per GIFT:

> Do the markers observed in this genome support at least one complete curated
> implementation of this capability?

It does not answer whether the capability is expressed, active, used, or
consequential. Keep that separation in mind from the start, because every
number later in these tutorials inherits it.

## Step 1: get your annotations into three columns

gifter takes a plain data frame. Whatever tool produced your annotations —
kofamscan, dbCAN, eggNOG-mapper, DRAM, bakta — the job is to reshape its output
into one row per observed marker:

| column | meaning |
|---|---|
| `gene_id` | your identifier for the gene. Optional, but it is what makes a call traceable back to a locus |
| `namespace` | which accession system: `KO`, `EC`, `PFAM`, `TIGRFAM`, `CAZY` |
| `accession` | the accession itself |

Long form matters: a gene hit by both a KO and a CAZy family is **two rows**,
not one row with two columns.

Here is an illustrative genome. It stands for an anaerobic gut bacterium that
degrades xylan, and it is a fixture chosen to exercise the model rather than
annotation output from a sequenced organism.

```{r}
gut_anaerobe <- data.frame(
  gene_id = c(
    # xylan backbone cleavage and debranching
    "g0101", "g0102", "g0103",
    # xylose transport and catabolism
    "g0210", "g0211", "g0212", "g0330", "g0331",
    # purine biosynthesis
    "g0450", "g0451", "g0452", "g0453", "g0454", "g0455", "g0456",
    # chemotaxis
    "g0620", "g0621", "g0622", "g0623", "g0624", "g0625", "g0626",
    # restriction-modification
    "g0810", "g0811", "g0812",
    # amino acid biosynthesis
    "g0910", "g0911", "g0912", "g0913", "g0914",
    # housekeeping genes gifter does not use
    "g1001", "g1002", "g1003"
  ),
  namespace = c(rep("CAZY", 3), rep("KO", 30)),
  accession = c(
    "GH11", "GH39", "CE1",
    "K10543", "K10544", "K10545", "K01805", "K00854",
    "K00764", "K01945", "K00601", "K01952", "K01587", "K01756", "K00602",
    "K03412", "K00575", "K03407", "K03408", "K03776", "K03413", "K03410",
    "K03427", "K01153", "K01154",
    "K00812", "K01915", "K00600", "K00826", "K00259",
    "K02945", "K02863", "K01873"
  ),
  stringsAsFactors = FALSE
)

nrow(gut_anaerobe)
```

If you only have a bare vector of accessions, gifter will infer the namespace
for KO, EC, Pfam, TIGRFAM and CAZy patterns. Supplying `namespace` explicitly
is safer.

## Step 2: see which markers gifter can use

Before evaluating anything, check how much of your annotation the reference
database actually recognises. `map_markers()` keeps every observation,
including the ones nothing matched, so marker use is auditable rather than
silent.

```{r}
mapped <- map_markers(gut_anaerobe)

# One row per (observation, curated component) pair, plus unmatched rows.
table(mapped$matched)
```

The unmatched markers are not a problem — a genome carries thousands of genes
and gifter curates a specific catalogue — but you should know which they are:

```{r}
unique(mapped$accession[!mapped$matched])
```

Those three are ribosomal and tRNA-synthetase orthologues. They are real genes;
gifter simply makes no capability claim that rests on them.

## Step 3: evaluate

```{r}
result <- evaluate_gifts(gut_anaerobe)
result
```

Every curated GIFT is evaluated, so the result always has the same number of
rows. That matters later: a genome's negative calls are part of the answer, not
an absence of one.

```{r}
nrow(result$gifts)
```

### The calls that came out complete

```{r}
complete <- result$gifts[result$gifts$complete, ]
complete[, c("gift_id", "gift_type", "evidence_confidence")]
```

Read the `gift_type` column. It tells you which completeness model produced the
call, and therefore what the call is allowed to mean:

- **metabolic** — at least one complete curated enzymatic route between the
  declared boundary molecules;
- **regulatory** — at least one complete curated circuit that senses a signal
  and executes a response;
- **defense** — at least one complete curated defense mechanism;
- **structural** — at least one complete curated architecture of a cellular
  structure.

Note `evidence_confidence` on `xylan_degradation`: **ambiguous**. The call rests
on CAZy family-level accessions, and a glycoside hydrolase family is
polyspecific. gifter reports the call at the confidence its weakest accepted
marker allows rather than presenting it as equal to a call built on
characterised orthologues.

## Step 4: read the incomplete calls, because they are informative

An incomplete call is not a blank. gifter reports the **closest** implementation
and exactly what it was missing, so you can tell a capability that is one step
short from one that is entirely absent.

```{r}
gifts <- result$gifts
near <- gifts[!gifts$complete & gifts$minimum_missing_requirements == 1, ]
near <- near[order(-near$completeness_score), ]
head(near[, c("gift_id", "gift_type", "completeness_score")], 5)
```

Take the top one:

```{r}
purine <- gifts[gifts$gift_id == "purine_core_biosynthesis", ]
purine$best_route
purine$minimum_missing_reactions
unlist(purine$missing_reactions_best_route)
length(unlist(purine$supporting_reactions))
```

Nine of the ten reactions of the closest route are supported and one is not.
That is a specific, checkable claim — go and look at whether `RHEA:23032` is
genuinely absent or whether your annotation missed it — and it is a very
different situation from a capability with no evidence at all.

Note what gifter did **not** do: it did not call the GIFT complete because 90% of
its genes were present. Completeness comes from route logic. The score is a
secondary summary, never the decision.

## Step 5: the specificity lesson

Look at these two regulatory calls together:

```{r}
regulatory <- gifts[gifts$gift_type == "regulatory", ]
regulatory[, c("gift_id", "complete", "minimum_missing_requirements")]
```

The genome completes `chemotaxis_signal_transduction` but not
`aspartate_chemoreception`. What is missing from the second?

```{r}
aspartate <- gifts[gifts$gift_id == "aspartate_chemoreception", ]
unlist(aspartate$missing_requirements)
```

The genome carries a generic methyl-accepting chemoreceptor. That accession is
assigned to dozens of receptors in a single genome, so it supports the claim
"this organism has a chemosensory pathway" and cannot support "this organism
senses aspartate". A curated marker anchored on the characterised aspartate
receptor would be needed for the second.

This is a rule, not an accident: **the specificity of a claim never exceeds the
specificity of its evidence.** Widening the generic marker to make the specific
GIFT fire would damage every other trait that marker also matches.

The same logic runs the other way between types:

```{r}
gifts[gifts$gift_id == "flagellar_apparatus",
      c("complete", "minimum_missing_requirements")]
```

This genome has a complete chemosensory pathway and no flagellum. Fusing the
two into a "motility" trait would have hidden exactly that.

## Step 6: trace a call back to genes

This is the step that makes a call reviewable. `trace_gift()` walks the whole
evidence chain down to your own gene identifiers.

```{r}
trace <- trace_gift(result, "xylose_degradation_isomerase")
trace[, c("reaction_id", "system_id", "component_id", "accession", "gene_id")]
```

The chain for a metabolic GIFT runs route → reaction → enzyme system →
component → marker → gene. For a machinery GIFT it runs implementation →
function → system → component → marker → gene, under that type's own names:

```{r}
defense_trace <- trace_gift(result, "type_i_restriction_modification")
head(defense_trace[, c("function_id", "system_id", "accession", "gene_id")])
```

If a reviewer disagrees with a call, this is where the disagreement becomes
concrete: a specific marker accepted for a specific component.

## Step 7: how the calls connect

Metabolic GIFTs compose through their declared boundary molecules — anchors —
and nothing else. Internal reaction intermediates never create a connection.

```{r}
edges <- gift_graph()
touching <- edges$from_gift %in% complete$gift_id |
            edges$to_gift %in% complete$gift_id
edges[touching, c("from_gift", "shared_anchor", "to_gift", "edge_quality")]
```

`edge_quality` is worth reading. `exact` means both GIFTs declared the same
anchor. `compartment_inexact` means they declared the same molecule but one
side left its compartment unresolved, so the link assumes a location curation
did not license.

Circular metabolism is derived from this graph rather than curated:

```{r}
evaluate_gift_cycles(result)
```

`status` is assembled from calls that were already made. Whether a genome closes
a cycle never changes whether any member GIFT is complete.

## What a positive call still does not mean

Worth restating now that you have some:

- it does not mean the machinery is expressed or the pathway carries flux;
- it does not mean the substrate is available in the organism's environment;
- `type_i_restriction_modification` does not mean the genome is phage-resistant;
- `chemotaxis_signal_transduction` does not mean the organism is motile — this
  genome is the proof, since it has no flagellum.

## Where to go next

Vignette 2, *From calls to quantitative traits*, turns a set of calls into
numbers that can be compared between genomes — and shows why every such number
needs a stated denominator. Vignette 3, *A genome-resolved community*, does the
same across many genomes at once.
