---
title: "4. Many samples over one catalogue"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4. Many samples over one catalogue}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Vignette 3 read one community. Real work is rarely one community: you assemble a
MAG catalogue once, map every sample against it, and end up with a genome by
sample abundance table in which different genomes are detected, at different
abundances, in each sample.

Binding one community per sample would be wrong twice over. It would evaluate
the same genomes once per sample for calls that cannot differ, and it would let
the samples end up with inconsistent call sets — the failure the release check
in `gifter_community()` already guards against one level up.

```{r}
library(gifter)
```

## Calls are a property of a genome

This is the whole design. `evaluate_gifts()` reads markers; it has never seen a
sample and cannot. Across a dataset exactly two things vary: **which genomes are
members** of a sample's community, and **how much of that sample** each one
represents. So the catalogue is evaluated once, and a sample is a restriction and
a reweighting of that one reading.

The same four illustrative genomes as vignette 3, around the curated
arabinoxylan chain.

```{r}
markers <- function(namespace, accession) {
  data.frame(
    gene_id = paste0("gene_", seq_along(accession)),
    namespace = namespace, accession = accession,
    stringsAsFactors = FALSE
  )
}

# A strips the side chains; B cleaves the backbone and releases xylose outside
# the cell; C and D take that xylose up and catabolise it.
catalogue <- gifter_community(
  A = evaluate_gifts(markers("KO", "K01209")),
  B = evaluate_gifts(markers("CAZY", c("GH11", "GH39"))),
  C = evaluate_gifts(markers("KO", c("K10543", "K10544", "K10545", "K01805", "K00854"))),
  D = evaluate_gifts(markers("KO", c("K10543", "K10544", "K10545", "K01805", "K00854")))
)
```

## The dataset

A genome by sample abundance matrix, carrying both sets of identifiers.

```{r}
abundance <- matrix(
  c(
    0.40, 0.30, 0.20, 0.10,   # s1: everything present
    0.60, 0.40, 0.00, 0.00,   # s2: no consumers
    0.00, 0.50, 0.40, 0.10    # s3: no debrancher
  ),
  nrow = 4,
  dimnames = list(c("A", "B", "C", "D"), c("s1", "s2", "s3"))
)

metadata <- data.frame(
  sample_id = c("s1", "s2", "s3"),
  group = c("fibre", "control", "fibre"),
  stringsAsFactors = FALSE
)

dataset <- gifter_dataset(catalogue, abundance, metadata)
dataset
```

A long data frame with `sample_id`, `genome_id` and `abundance` columns — the
shape coverM and its relatives emit — is accepted too and pivots to the same
matrix. It has to carry a row for every genome in every sample: after the pivot
an absent row and an abundance of zero are indistinguishable, and a row dropped
by a failed join would silently take an abundant genome out of a sample's
community.

The identifiers are not optional anywhere. A matrix without dimnames, a genome
in the catalogue that the abundance does not name, a metadata table that does not
cover the samples exactly — each is an error rather than a silent zero, because
each is far more often a failed join than a biological result.

Any sample can still be read by every function you already know:

```{r}
sample_community(dataset, "s2")
```

## Detection is to samples what assessability is to genomes

Both may only move denominators. Assessability decides whether a genome's silence
about a GIFT is informative; detection decides whether a genome is part of a
sample's community at all. Neither can promote an unsupported GIFT to supported,
and neither changes a call.

That is why `detection` is an argument of the *readers*, not of the container —
the same reasoning that puts `threshold` on `community_traits()` rather than on
`gifter_community()`. One dataset, read at two thresholds, without rebuilding
anything.

```{r}
sample_community(dataset, "s1")$genome_id
sample_community(dataset, "s1", detection = 0.15)$genome_id
```

## Per-sample traits

```{r}
universe <- gift_universe(label = "all curated GIFTs")
traits <- dataset_traits(dataset, universes = list(universe), progress = FALSE)
traits
```

Every sample-level richness is reported beside `detected_genomes`, because a
genome absent from a sample may be below detection rather than genuinely absent
and gifter models no sequencing depth. Nothing here imputes one, corrects for
one, or rarefies; the honest thing to do is to report the denominator.

```{r}
rows <- traits$metrics[
  traits$metrics$metric_id %in% c("community_richness", "detected_genomes"), ]
rows[, c("sample_id", "metric_id", "value", "numerator", "denominator")]
```

Per GIFT, presence and abundance stay in separate rows and never merge. How many
genomes encode a capability and how much of the sample's abundance those genomes
carry are different questions, and neither is a statement about activity.

```{r}
uptake <- traits$metrics[traits$metrics$target_id == "xylose_uptake_abc", ]
uptake[, c("sample_id", "metric_id", "value", "numerator", "denominator")]
```

Abundance is closed within each sample's **detected** set, after detection is
applied, so `abundance_coverage` is a share of the community actually being
described. Above `detection = 0` that differs from closing over the whole
catalogue — deliberately, since a genome the reader has just excluded should not
sit in the denominator.

### Two tables, because some rows have no sample

A genome supports the same GIFTs wherever it is detected, and two genomes share
the same repertoire wherever both are. Those rows cannot vary between samples,
so they are not given a `sample_id` column that would always be empty:

```{r}
head(traits$catalogue_metrics[, c("target_type", "target_id", "metric_id", "value")])
```

The trace works the same way. A sample's providers are the catalogue's providers
intersected with its detected genomes, so the per-sample rows are derived on
demand rather than stored once per sample:

```{r}
trace <- trace_sample(traits, "s3")
trace[trace$metric_id == "provider_count", c("target_id", "contribution")]
```

## Topology per sample

`dataset_network()` builds the catalogue's edge set once and restricts it to each
sample's detected genomes. No compatibility rule is invented and none is relaxed:
a cross-genome edge still requires the producing GIFT's output anchor to be
declared `extracellular`, and every edge still carries the `edge_quality` of the
GIFT edge beneath it.

```{r}
network <- dataset_network(dataset)
network$edges[, c("sample_id", "from_genome", "to_genome", "shared_anchor")]
```

The chain is complete in `s1`. In `s2` nobody consumes the released xylose, and
in `s3` nobody exposes the backbone — which is a difference in who is present,
not a difference in what anybody encodes.

## Where gifter stops

gifter computes per-sample traits and emits them joined to your metadata. It runs
**no** hypothesis test, differential-abundance analysis, ordination or effect
size between groups of samples, and it interprets no metadata column.

That is a deliberate boundary rather than an omission. A group label is a
user-supplied vector multiplied through the call matrix; presenting the result as
a gifter inference would put your design, your contrasts and your
multiple-testing correction under gifter's name. vegan, lme4, MaAsLin and ALDEx2
already do that work, and do it better.

What they cannot get anywhere else is an **assessability-aware matrix with a
declared reference universe**:

```{r}
capabilities <- gift_matrix(
  dataset,
  quality = c(A = 0.98, B = 0.96, C = 0.45, D = 0.97),
  policy = "completeness", threshold = 0.9
)

dim(capabilities)
attr(capabilities, "reference_universe")
attr(capabilities, "database_version")
table(capabilities, useNA = "ifany")
```

Rows are genomes, columns are GIFTs, and the third state is the point. C is too
fragmented for its silence to be read, so its negative calls are `NA` rather than
`0`. A zero there is a fabricated absence, and every model fitted on it inherits
the fabrication with nothing downstream able to see it.

## Handing it over

One metric, reshaped to samples by target, is what a distance function wants:

```{r}
coverage <- dataset_matrix(traits, "abundance_coverage", fill = 0)
round(coverage, 3)
```

`fill = 0` is your claim, not gifter's: a cell with no row may be a true zero or
a GIFT nothing in that sample could assess, so the default is `NA` and gifter
does not guess.

From here the analysis leaves gifter. With `vegan` installed, the two lines
below are where the boundary actually falls:

```{r, eval = FALSE}
distances <- vegan::vegdist(coverage, method = "bray")
vegan::adonis2(distances ~ group, data = metadata)
```

The design is yours, the test is `vegan`'s, and the matrix is gifter's — carrying
its reference universe and its database version so that the columns mean the same
thing in six months.

The tidy table joins the per-sample metrics to your metadata for the same
purpose:

```{r}
frame <- as.data.frame(traits)
head(frame[frame$metric_id == "community_richness",
           c("sample_id", "group", "value", "assessable")])
```

## Putting the four tutorials together

```text
annotation table
      |  evaluate_gifts_community()
      v
  GIFT calls
      |
      |  gifter_community()                 one catalogue, evaluated once
      v
 the catalogue -----> community_traits(), community_network()   (one sample)
      |
      |  + abundance (genome x sample), + metadata
      v
 gifter_dataset ----> dataset_traits(), dataset_network()       (many samples)
      |
      |  gift_matrix(), dataset_matrix(), as.data.frame()
      v
   vegan, lme4, MaAsLin, ALDEx2                                 (not gifter)
```

The discipline is the same at every step, and the last arrow is part of it. A
number carries the set it was counted over, the evidence that produced it stays
attached, and nothing becomes more certain than the calls underneath. A GIFT call
is what a genome encodes — not what an organism is doing, and not what
distinguishes one group of samples from another.
