---
title: "2. From calls to quantitative traits"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. From calls to quantitative traits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Vignette 1 produced 130 calls for one genome. This one turns them into numbers
you can compare between genomes — and shows why a number of that kind is
meaningless until you say what it was counted over.

```{r}
library(gifter)
```

We reuse the illustrative gut anaerobe from vignette 1.

```{r}
gut_anaerobe <- data.frame(
  gene_id = paste0("g", sprintf("%04d", 1:33)),
  namespace = c(rep("CAZY", 3), rep("KO", 30)),
  accession = c(
    "GH11", "GH39", "CE1",
    "K10543", "K10544", "K10545", "K01805", "K00854",
    "K00764", "K01945", "K00601", "K01952", "K01587", "K01756", "K00602",
    "K03412", "K00575", "K03407", "K03408", "K03776", "K03413", "K03410",
    "K03427", "K01153", "K01154",
    "K00812", "K01915", "K00600", "K00826", "K00259",
    "K02945", "K02863", "K01873"
  ),
  stringsAsFactors = FALSE
)

result <- evaluate_gifts(gut_anaerobe)
sum(result$gifts$complete)
```

## The problem: five is not a result

This genome supports ten GIFTs. Is that a lot?

The question has no answer until you say ten *out of what*. Out of the 130
GIFTs curated today? Out of the 122 metabolic ones? Out of the amino-acid
biosynthetic capabilities, of which this genome was tested for none? And when
the next curation campaign adds forty GIFTs, does "ten" still mean the same
thing?

gifter's answer is that every quantitative trait names its **reference
universe**, and carries that name in the output beside the number.

## Step 1: define a universe

`gift_universe()` builds one from curated metadata — the GIFT type, the
metabolic mode, the registered facet vocabulary, the derived profile view. You
never write a list of GIFT identifiers yourself; that would move biological
content out of the database, where it can be validated and versioned, and into
your script, where it cannot.

Common questions already have named, versioned presets. Inspect them before
writing a custom filter:

```{r}
presets <- list_gift_universes()
presets[, c("universe_id", "bounded", "member_count")]

carbohydrate <- gift_universe(preset = "carbohydrate_degradation")
carbohydrate
```

### Choose a preset from your research question

The searchable [Reference universes atlas](https://alberdilab.github.io/gifter/atlas/#universes)
is the quickest chooser: search for a substrate, process or metric, then narrow
the cards to genome, community, network or bounded-coverage analyses. The full
registry is reproduced below so that the choice remains visible in this
tutorial too. "Current GIFTs" describes this database release, not a fixed list
stored in the preset.

```{r universe-guide, echo = FALSE}
guide <- presets[, c(
  "label", "universe_id", "description", "member_count",
  "recommended_metrics", "bounded"
)]
guide$bounded <- ifelse(
  guide$bounded,
  "Bounded: coverage fractions are valid",
  "Open: use counts, breadth or distribution"
)
guide$recommended_metrics <- gsub(
  " \\| ", ", ", guide$recommended_metrics
)
names(guide) <- c(
  "Research question", "Preset", "Use it for", "Current GIFTs",
  "Recommended starting metrics", "Denominator"
)
knitr::kable(guide, row.names = FALSE)
```

Choose the universe from the biological question first, then use its
scope-specific metric recommendations:

- for one genome, start with `gift_richness` or the recommended breadth metric;
- for a community, use `community_richness` to count represented capabilities
  and `provider_count` for redundancy, keeping abundance separate;
- use a network metric only when the question depends on compatible,
  extracellular GIFT boundaries;
- use `supported_fraction` or `community_coverage` only where the table marks
  the universe bounded, and always report `assessable_fraction` alongside a
  genome fraction.

Each preset carries its description, interpretation limits and recommended
genome, community or network metrics. Its members are still resolved from the
current release's metadata: the registry stores a query, never a list of GIFT
identifiers. An open preset cannot be promoted to `bounded = TRUE` at runtime.

For a question that has no preset, build a custom universe directly:

```{r}
gift_universe(type = "metabolic")

gift_universe(mode = "catabolic")

gift_universe(facet = "substrate_class", value = "polysaccharide")
```

Filters combine with AND:

```{r}
gift_universe(mode = "catabolic", resource_strategy = "uptake")
```

Note that `mode` partitions this database far better than `gift_type` does —
122 of the 130 GIFTs are metabolic, so "metabolic GIFTs" is nearly the whole
catalogue, while anabolic/catabolic/transport/interconversion splits it 57 /
58 / 3 / 4.

## Step 2: compute traits

```{r}
traits <- genome_traits(result, genome_id = "MAG_001")
traits
```

With no universes supplied you get a default set that partitions the catalogue
by type, by mode, and by resource strategy. The result is a long-form table:

```{r}
names(traits$metrics)
```

Every row carries its numerator, its denominator, how many members of the
universe were assessable, the universe itself, and the database version. That
is what lets a number be taken apart six months later.

```{r}
richness <- traits$metrics[traits$metrics$metric_id == "gift_richness", ]
richness[, c("reference_universe", "value", "assessable")]
```

## Step 3: ask why a number is what it is

Every metric keeps the GIFTs that produced it.

```{r}
trace <- traits$trace[
  traits$trace$metric_id == "gift_richness" &
    traits$trace$reference_universe == "catabolic GIFTs",
]
trace[, c("metric_id", "gift_id")]
```

This is the point of the long-form shape. A trait you cannot decompose is a
trait you cannot defend.

## Step 4: breadth is not richness

Counting GIFTs overcounts wherever the ontology happens to be finely
subdivided. Breadth counts distinct classifications instead:

```{r}
breadth <- traits$metrics[
  startsWith(traits$metrics$metric_id, "breadth_") &
    traits$metrics$reference_universe == "all curated GIFTs",
]
breadth[, c("metric_id", "value", "denominator")]
```

`breadth_substrate_class` counts how many curated substrate classes this genome
can act on, out of the classes present in the universe. Two GIFTs acting on the
same class contribute one class, not two.

## Step 5: fractions, and when gifter refuses to give you one

Here is a metric that is **not** in the output above:

```{r}
"supported_fraction" %in% richness$metric_id
```

gifter will not tell you what fraction of the metabolic catalogue this genome
supports. Supporting 8 of 122 does not mean a genome lacks 114 capabilities; it
means the catalogue currently stops at 122 and will not tomorrow. A fraction of
an open, growing set reads as "the share of microbial function this organism
carries", which is not what it measures.

A fraction is offered only for a universe you explicitly declare **bounded** —
a claim that curation intends to cover that set completely.

```{r}
autonomy <- gift_universe(preset = "biomass_essential_anabolism")
autonomy

autonomy_traits <- genome_traits(
  result, universes = list(autonomy), genome_id = "MAG_001"
)
autonomy_traits$metrics[
  autonomy_traits$metrics$metric_id %in%
    c("gift_richness", "supported_fraction"),
  c("metric_id", "value", "numerator", "denominator")
]
```

Over that universe the fraction *is* meaningful: it is genomic coverage of the
curated biomass-essential anabolic capabilities. It still does not mean the
organism grows without supplementation — it means the genes for those
capabilities are or are not there.

## Step 6: MAG completeness, and the honesty of a denominator

Everything above assumed that a negative call means absence. For a closed
isolate genome that is reasonable. For a 55%-complete metagenome-assembled
genome it is not: the organism was never fully observed, so it has not been
shown to lack anything.

By default gifter makes the optimistic assumption and *tells you it did*:

```{r}
assessable <- traits$metrics[
  traits$metrics$metric_id == "assessable_fraction" &
    traits$metrics$reference_universe == "all curated GIFTs",
]
assessable[, c("value", "numerator", "denominator")]
```

Supply a completeness estimate and a threshold, and negative calls on a genome
below that threshold leave every denominator. Both may be stated as proportions
or as the percentages CheckM and BUSCO report — a proportion cannot exceed 1, so
gifter reads the scale from the values themselves, and `quality = c(MAG_001 =
55), threshold = 90` is the same analysis as the call below:

```{r}
fragmented <- genome_traits(
  result, universes = list(autonomy), genome_id = "MAG_001",
  quality = c(MAG_001 = 0.55), policy = "completeness", threshold = 0.9
)
fragmented$metrics[
  fragmented$metrics$metric_id %in%
    c("gift_richness", "supported_fraction", "assessable_fraction"),
  c("metric_id", "value", "numerator", "denominator")
]
```

Read that carefully, because it is the most misusable output in the package.
`supported_fraction` went from 0.125 to **1.0**, and the genome did not gain a
single capability: `gift_richness` is 5 in both runs. All that happened is that
the 35 negative calls left the denominator, because a 55%-complete genome has
not been shown to lack them. The number now means "of the five capabilities we
could assess, five are supported", which is nearly content-free.
`assessable_fraction` is what says so — 5 of 40 — and the warning gifter emitted
points at exactly that.

The honest reading of a fragmented genome is `gift_richness` beside
`assessable_fraction`, not the fraction on its own.

Two guarantees hold whatever policy you choose:

- **no policy can make an unsupported GIFT supported.** Genome quality changes
  how absence is read and nothing else. The complete calls are identical in
  both runs above.
- there is **no default threshold**. How complete a genome must be before its
  silence is informative is your declared choice, and a package default would be
  read as a recommendation.

## What these numbers may not say

- richness counts encoded capabilities in the current gifter ontology within a
  stated universe. It is not a measure of biological complexity or of metabolic
  versatility in any particular environment.
- a supported fraction is genomic coverage, never growth independence,
  activity, flux or phenotype.
- nothing here has become more certain than the calls underneath it. A trait
  built on an `ambiguous` call inherits that ambiguity.

## Where to go next

Vignette 3, *A genome-resolved community*, applies the same discipline across
many genomes: how redundantly a capability is provided, which genome is its only
provider, and which chains the community completes that no single member does.
