---
title: "Across chicken caecal maturation"
description: >
  Encoded capabilities in 822 bacterial MAGs across 388 samples from two
  chicken trials, with detection and evidence sensitivity kept explicit.
---

```{r setup, include = FALSE}
# This article reads committed analysis products. It does not rerun annotation,
# evaluation, detection, network construction or statistical models.
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  dev = "ragg_png", dpi = 150, fig.retina = 2, out.width = "100%",
  fig.width = 9, fig.height = 5.4, fig.align = "center"
)
default_plot_hook <- knitr::knit_hooks$get("plot")
knitr::knit_hooks$set(plot = function(x, options) {
  href <- paste0("chicken_files/figure-html/", basename(x[[1L]]))
  paste0(
    '<a class="case-figure" href="', href, '" target="_blank" ',
    'title="Open at full resolution">', default_plot_hook(x, options), "</a>"
  )
})

library(ggplot2)

root <- normalizePath(".")
while (!file.exists(file.path(root, "DESCRIPTION"))) root <- dirname(root)
out <- file.path(root, "manuscript", "analysis", "output")
read_tsv <- function(file) {
  utils::read.delim(file.path(out, file), check.names = FALSE,
                    stringsAsFactors = FALSE, quote = "")
}
read_xz <- function(file) {
  utils::read.delim(xzfile(file.path(out, file)), check.names = FALSE,
                    stringsAsFactors = FALSE, quote = "")
}
item <- function(table, key) {
  value <- table$value[table$item == key]
  stopifnot(length(value) == 1L)
  value
}
one <- function(table, ...) {
  filters <- list(...)
  for (key in names(filters)) table <- table[table[[key]] == filters[[key]], , drop = FALSE]
  stopifnot(nrow(table) == 1L)
  table
}
fmt <- function(x, digits = 2) format(round(as.numeric(x), digits), nsmall = digits, trim = TRUE)
big <- function(x) format(as.numeric(x), big.mark = ",", trim = TRUE)

audit <- read_tsv("r10-input-audit.tsv")
followup <- read_tsv("r10-followup-audit.tsv")
frame_audit <- read_tsv("r10-frame-time-audit.tsv")
samples <- read_xz("r10-sample-traits.tsv.xz")
network <- read_tsv("r10-network-metrics.tsv")
age_contrasts <- read_tsv("r10-age-contrasts.tsv")
frames <- read_tsv("r10-frame-time-classification.tsv")
weighted <- read_tsv("r10-frame-time-abundance-weighted-classification.tsv")
mode <- read_tsv("r10-richness-mode-samples.tsv")
driver <- read_tsv("r10-richness-mode-driver.tsv")
mode_age <- read_tsv("r10-richness-mode-driver-age-summary.tsv")
handoff_graph <- read_tsv("r10-handoff-graph.tsv")
handoff_age <- read_tsv("r10-handoff-age-summary.tsv")
handoff_contrasts <- read_tsv("r10-handoff-age-contrasts.tsv")
size_audit <- read_tsv("r10-size-expectation-audit.tsv")
size_effects <- read_tsv("r10-size-expectation-age-contrasts.tsv")

primary <- as.numeric(item(audit, "primary_detection"))
stopifnot(primary == 0.001, nrow(mode) == as.numeric(item(audit, "samples")))

sample_median <- function(age, frame, metric) {
  x <- samples$value[samples$detection == primary &
                     samples$sampling_time == age &
                     samples$reference_frame == frame &
                     samples$metric_id == metric]
  stopifnot(length(x) > 0L)
  median(x)
}
handoff_median <- function(age, metric) {
  one(handoff_age, confidence_floor = "all accepted", detection = primary,
      sampling_time = age, metric_id = metric)$median
}

ink <- "#173330"; muted <- "#687672"; line <- "#dcded5"; surface <- "#fffdf8"
forest <- "#0d3d38"; mint <- "#90c9b5"; coral <- "#ef765d"; gold <- "#d9a936"
blue <- "#477e91"
ages <- c("day 7", "day 21", "day 35")
age_colour <- c("day 7" = forest, "day 21" = blue, "day 35" = coral)
theme_case <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      text = element_text(colour = ink),
      axis.text = element_text(colour = muted),
      axis.title = element_text(colour = ink, size = rel(0.92)),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = line, linewidth = 0.35),
      plot.background = element_rect(fill = surface, colour = NA),
      panel.background = element_rect(fill = surface, colour = NA),
      legend.background = element_rect(fill = surface, colour = NA),
      legend.key = element_rect(fill = surface, colour = NA),
      legend.position = "bottom",
      strip.text = element_text(colour = ink, face = "bold"),
      plot.margin = margin(10, 14, 8, 10)
    )
}
```

<p class="lead">What changes as a chicken caecal community matures: the union
of capabilities its members encode, or the repertoire carried by each member?
We followed `r item(audit, "published_bacterial_mags")` bacterial metagenome-assembled
genomes (MAGs) across `r item(audit, "samples")` samples at days 7, 21 and 35.
The two views tell different stories.</p>

<div class="case-stats">
<div><strong>`r item(audit, "published_bacterial_mags")`</strong><span>bacterial MAGs</span></div>
<div><strong>`r item(audit, "samples")`</strong><span>caecal samples</span></div>
<div><strong>`r item(audit, "catalogue_gifts")`</strong><span>GIFTs evaluated</span></div>
<div><strong>`r big(item(audit, "supported_genome_gift_calls"))`</strong><span>supported MAG–GIFT calls</span></div>
<div><strong>`r nrow(handoff_graph)`</strong><span>exact extracellular links</span></div>
</div>

**In short**

- Median community metabolic richness was `r sample_median(7, "All metabolic GIFTs", "community_richness")`,
  `r sample_median(21, "All metabolic GIFTs", "community_richness")` and
  `r sample_median(35, "All metabolic GIFTs", "community_richness")` GIFTs at the three ages.
- Median mean metabolic repertoire per detected MAG fell from
  `r fmt(sample_median(7, "All metabolic GIFTs", "mean_genome_richness"))` to
  `r fmt(sample_median(35, "All metabolic GIFTs", "mean_genome_richness"))` GIFTs.
- Three of the `r item(frame_audit, "curated_reference_frames")` curated
  reference frames had a decrease beyond one GIFT in per-MAG richness at both
  later ages. Their community unions remained within that margin.
- An apparent two-band pattern in community richness follows the operational
  detection of one broad-repertoire *Escherichia coli* MAG.

These are encoded capabilities. They say nothing directly about expression,
flux, nutrient exchange, growth or host phenotype. If a GIFT call is new to
you, [start with a single genome](evaluating-a-genome.html).

## The dataset and its boundary

We used the [public archive](https://doi.org/10.5281/zenodo.18457570) of the
[Marcos et al. chicken caecal study](https://doi.org/10.1093/ismeco/ycag091).
The archive's abundance matrix has 825 rows. An exact join to its published
bacterial taxonomy retained 822 MAGs. Their assemblies were checksum-verified
and reannotated with Drakkar's gene-resolved marker projection, producing
`r big(item(audit, "drakkar_marker_rows"))` marker rows. gifter evaluated those
markers into `r big(item(audit, "supported_genome_gift_calls"))` supported
MAG–GIFT pairs, across `r item(audit, "represented_gifts")` of the
`r item(audit, "catalogue_gifts")` catalogue GIFTs. This reading uses database
version `r unique(samples$database_version)`.

```{r evaluation-example, eval = FALSE, echo = TRUE}
library(gifter)

# Gene-resolved marker rows have genome_id, gene_id, namespace and accession.
catalogue <- evaluate_gifts_community(markers)
dataset <- gifter_dataset(catalogue, abundance, metadata)
```

The archived data lack the per-sample genome-breadth matrix used in the
source paper. We therefore call a MAG *detected* when its reconstructed
relative abundance is **greater than 0.001** for the primary reading, and
repeat the analysis at greater than 0, 0.00001 and 0.0001. This threshold is
an operational choice; it does not reproduce the source paper's breadth
filter. A completeness estimate below 90% makes an unsupported call
indeterminate for absence denominators only. Detection and completeness never
promote a GIFT call to supported.

## A broad union, narrower members

```{r age-data}
select_metric <- function(frame, metric, label, table = samples) {
  x <- table[table$detection == primary & table$reference_frame == frame &
             table$metric_id == metric, c("sampling_time", "trial", "value")]
  x$panel <- label
  x
}
main <- rbind(
  select_metric("All metabolic GIFTs", "community_richness", "Community metabolic richness"),
  select_metric("All metabolic GIFTs", "mean_genome_richness", "Mean per-MAG metabolic repertoire"),
  select_metric("Plant fibre utilisation", "community_richness", "Community plant-fibre richness"),
  select_metric("Plant fibre utilisation", "interaction_density", "Exact potential-handoff density", network)
)
main$sampling_time <- factor(paste("day", main$sampling_time), levels = ages)
main$panel <- factor(main$panel, levels = c(
  "Community metabolic richness", "Mean per-MAG metabolic repertoire",
  "Community plant-fibre richness", "Exact potential-handoff density"
))
```

```{r age, fig.width = 9.5, fig.height = 7, fig.cap = "Sample-level results at relative abundance > 0.001. Boxes show distributions within each age and trial; points are samples. Density counts compatible ordered MAG pairs over eligible ordered pairs for the exact plant-fibre network. It is potential topology, not observed exchange."}
ggplot(main, aes(sampling_time, value, colour = trial)) +
  geom_boxplot(aes(group = interaction(sampling_time, trial)),
               outlier.shape = NA, width = 0.62, alpha = 0.18) +
  geom_jitter(width = 0.16, height = 0, alpha = 0.34, size = 0.8) +
  facet_wrap(~ panel, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = c(CA = forest, CB = coral)) +
  labs(x = NULL, y = NULL, colour = "Trial") + theme_case()
```

Median detected MAG count rose from
`r sample_median(7, "All metabolic GIFTs", "detected_genomes")` at day 7 to
`r sample_median(35, "All metabolic GIFTs", "detected_genomes")` at day 35.
Community metabolic richness remained near 110, while mean per-MAG metabolic
richness fell. In models adjusting for trial, treatment, breed and sex, with a
random intercept for pen, the day-21 and day-35 differences in mean per-MAG
richness from day 7 were
`r fmt(one(age_contrasts, detection = primary, reference_frame = "All metabolic GIFTs", metric_id = "mean_genome_richness", contrast = "day_21_minus_day_7")$estimate)` and
`r fmt(one(age_contrasts, detection = primary, reference_frame = "All metabolic GIFTs", metric_id = "mean_genome_richness", contrast = "day_35_minus_day_7")$estimate)`
GIFTs. Those directions persisted at every tested detection threshold. The
community's bounded biomass-essential coverage stayed at a median of
`r fmt(100 * sample_median(7, "biomass-essential anabolic GIFTs", "community_coverage"), 1)`%
at all three ages. A broad community union can therefore coexist with a
narrower encoded repertoire in each detected member.

## What the curated frames add

Each [reference frame](quantitative-traits.html) is a database-defined set
of GIFTs. The analysis compared all `r item(frame_audit, "curated_reference_frames")`
frames. A change was called *stable within margin* only when both later-age
contrasts were equivalent within one GIFT for counts, or five percentage
points for bounded coverage, at **all four** detection thresholds. A
nonsignificant age test by itself was not treated as evidence of stability.

```{r frame-data}
tile <- frames[frames$metric_id %in% c("community_richness", "mean_genome_richness"), ]
tile$reading <- paste(
  ifelse(tile$metric_id == "community_richness", "Community", "Per MAG"),
  ifelse(tile$contrast == "day_21_minus_day_7", "day 21", "day 35")
)
tile$reading <- factor(tile$reading, levels = c(
  "Community day 21", "Community day 35", "Per MAG day 21", "Per MAG day 35"
))
order_frames <- tile$reference_frame[tile$reading == "Per MAG day 35"]
tile$reference_frame <- factor(tile$reference_frame, levels = rev(order_frames))
labels <- c(
  observed_invariant = "Invariant", stable_within_margin = "Within margin",
  decrease = "Decrease > 1 GIFT", magnitude_uncertain = "Magnitude uncertain",
  detection_sensitive = "Detection sensitive"
)
```

```{r frames, fig.width = 9.5, fig.height = 8.3, fig.cap = "Classification of adjusted day-21 and day-35 changes from day 7 for all 19 curated frames. Community union and mean per-MAG richness are separate metrics. The colour describes a model and detection-sensitivity result, not a GIFT call or measured activity."}
ggplot(tile, aes(reading, reference_frame, fill = classification)) +
  geom_tile(colour = surface, linewidth = 0.8, width = 0.96, height = 0.9) +
  scale_fill_manual(values = c(
    observed_invariant = forest, stable_within_margin = mint,
    decrease = coral, magnitude_uncertain = gold,
    detection_sensitive = blue
  ), labels = labels, name = NULL) +
  labs(x = NULL, y = NULL) +
  theme_case(11) +
  theme(panel.grid = element_blank(), axis.text.x = element_text(angle = 25, hjust = 1),
        legend.position = "bottom")
```

Fifteen frames had community richness within the one-GIFT margin at every
threshold; amino-acid and nucleotide autonomy were exactly invariant.
Aromatic catabolism and carbon acquisition were detection sensitive. All
four bounded frames retained community coverage within five percentage
points. The clearest per-MAG decreases were in amino-acid autonomy,
biomass-essential anabolism and vitamin biosynthesis.

```{r focal-table}
focal_ids <- c("amino_acid_autonomy", "biomass_essential_anabolism", "vitamin_biosynthesis")
focal <- frames[frames$frame_id %in% focal_ids &
                frames$metric_id == "mean_genome_richness", ]
focal <- focal[order(match(focal$frame_id, focal_ids), focal$contrast), ]
weighted_focal <- weighted[weighted$frame_id %in% focal_ids, ]
stopifnot(nrow(weighted_focal) == nrow(focal), all(weighted_focal$estimate < 0))
knitr::kable(
  data.frame(
    Frame = focal$reference_frame,
    Contrast = ifelse(focal$contrast == "day_21_minus_day_7", "day 21 − day 7", "day 35 − day 7"),
    `Adjusted difference` = fmt(focal$estimate),
    `95% interval` = paste0("[", fmt(focal$confidence_low), ", ", fmt(focal$confidence_high), "]"),
    check.names = FALSE
  ),
  align = "llrr",
  caption = "Mean encoded GIFTs per detected MAG. Each of these six effects exceeds the one-GIFT margin; intervals and classifications come from the committed model output."
)
```

The abundance-weighted reading sums the relative abundance coverage of each
member GIFT within the detected community. It preserved all
`r nrow(weighted_focal)` negative age directions.
`r sum(weighted_focal$classification == "decrease")` effects stayed beyond one
GIFT; the day-21 vitamin result had an uncertain magnitude when weighted by
abundance. Weighting changes how an
encoded repertoire is summarised; it does not measure expression or flux.

## Why community richness forms two bands

At the primary threshold, `r driver$lower_mode_samples` samples held
`r min(mode$metabolic_richness[mode$richness_mode == "lower"])`–`r driver$mode_lower_maximum`
metabolic GIFTs and `r driver$upper_mode_samples` held
`r driver$mode_upper_minimum`–`r max(mode$metabolic_richness[mode$richness_mode == "upper"])`.
The detected status of `r driver$driver_genome`, an *`r driver$species`* MAG,
matched the split in all `r driver$samples` samples. It encoded
`r driver$metabolic_gifts_in_driver` metabolic and
`r driver$plant_fibre_gifts_in_driver` plant-fibre GIFTs.

```{r mode, fig.width = 9, fig.height = 5.5, fig.cap = "Each point is one sample. The dashed line is the operational detection cutoff of relative abundance > 0.001 for cmag_510. The apparent two-band richness pattern follows whether this broad-repertoire MAG lies above the cutoff; abundance remains continuous."}
mode$age <- factor(paste("day", mode$sampling_time), levels = ages)
ggplot(mode, aes(driver_abundance, metabolic_richness, colour = age)) +
  geom_vline(xintercept = primary, linetype = "dashed", colour = muted) +
  geom_point(alpha = 0.7, size = 1.8) +
  scale_x_log10() +
  scale_colour_manual(values = age_colour, name = NULL) +
  labs(x = "cmag_510 relative abundance (log scale)",
       y = "Community metabolic GIFT richness") + theme_case()
```

Computationally removing this MAG's calls reduced median metabolic richness
to `r median(mode$richness_without_driver_metabolic[mode$richness_mode == "lower"])`
in the lower band and
`r median(mode$richness_without_driver_metabolic[mode$richness_mode == "upper"])`
in the upper band. Its median abundance fell from
`r fmt(one(mode_age, sampling_time = 7)$median_abundance, 5)` at day 7 to
`r fmt(one(mode_age, sampling_time = 35)$median_abundance, 5)` at day 35,
moving more samples below the cutoff. At a threshold of 0.00001 it was
detected in every sample. The bands are thus a sensitivity of a binary
community union to detection, not two measured functional states.

## Exact potential handoffs

A cross-genome link requires two GIFTs to share a **declared extracellular
anchor** exactly. The catalogue at this database version yielded only
`r nrow(handoff_graph)` such links:

```{r handoff-table}
knitr::kable(
  data.frame(
    Provider = gsub("_", " ", handoff_graph$from_gift),
    `Extracellular anchor` = handoff_graph$shared_anchor,
    Recipient = gsub("_", " ", handoff_graph$to_gift),
    check.names = FALSE
  ), align = "lll",
  caption = "Exact graph links eligible for cross-genome projection. These are curated compatibility paths, not observed interactions."
)
```

Median potential-handoff density was
`r fmt(handoff_median(7, "interaction_density"), 3)`,
`r fmt(handoff_median(21, "interaction_density"), 3)` and
`r fmt(handoff_median(35, "interaction_density"), 3)` at the three ages.
The day-21 and day-35 adjusted density differences from day 7 were
`r fmt(one(handoff_contrasts, confidence_floor = "all accepted", detection = primary, metric_id = "interaction_density", contrast = "day_21_minus_day_7")$estimate, 3)` and
`r fmt(one(handoff_contrasts, confidence_floor = "all accepted", detection = primary, metric_id = "interaction_density", contrast = "day_35_minus_day_7")$estimate, 3)`.
The direction remained negative across detection thresholds and after
subtracting a null mean from
`r item(followup, "null_replicates_per_detected_genome_count")` random
catalogue communities matched for detected-MAG count. With ambiguous marker
evidence withheld, absolute density was much lower, but the age direction
did not reverse. Potential links are constrained by the three curated
extracellular paths; they do not show that molecules moved between organisms.

## A later comparison with genome size

The [phylogeny case study](phylogeny.html) provides a size expectation from
`r item(size_audit, "GTDB genomes behind the expectation")` complete genomes.
A later R10 analysis applied it to these MAGs using database
`r item(size_audit, "database version")` and
`r item(size_audit, "catalogue GIFTs")` GIFTs. This is a **separate
reevaluation** from the `r item(audit, "catalogue_gifts")`-GIFT result above;
the `r item(size_audit, "GIFTs shared with the audited R10 catalogue")`
shared GIFTs even have
`r item(size_audit, "calls on shared GIFTs that differ from the audited R10 calls")`
changed MAG calls. Its estimates must stay with its database release.

```{r size-effect}
size_day35 <- one(size_effects, variant = "primary", detection = primary,
                  frame_id = "all_metabolic", contrast = "day_35_minus_day_7")
```

In that later reading, the adjusted day-35 per-MAG metabolic difference from
day 7 was `r fmt(size_day35$observed)` GIFTs. The GTDB size expectation
reproduced `r fmt(size_day35$size_expected)` of that difference
(`r fmt(100 * size_day35$share_reproduced_by_size, 0)`%), leaving a
`r fmt(size_day35$deviation)`-GIFT deviation. Genome size thus explains part
of the narrower per-MAG repertoire, while the remaining difference stays
visible. It does not change any original call or turn a size association
into a functional mechanism.

## Reproduce this analysis

The [analysis directory](https://github.com/alberdilab/gifter/tree/main/manuscript/analysis)
contains the scripts, committed result tables and input manifests. The
[R10 case directory](https://github.com/alberdilab/gifter/tree/main/manuscript/analysis/r10-chicken)
pins the public archive, checksums, annotation transfer and detection limits.
Large marker inputs stay outside git; the scripts verify them before use.

| Script | What it contributes |
|---|---|
| `21-r10-chicken.R` | MAG calls, sample traits, age models and primary case figure |
| `22-r10-handoff-bimodality.R` | Richness-band diagnosis, exact extracellular links and matched null |
| `24-r10-reference-frame-time.R` | All curated frames, equivalence decisions and abundance weighting |
| `31-r10-size-expectation.R` | Separate current-database comparison with the GTDB size expectation |

Every plot here is drawn from the committed tables; the site build performs
no annotation, GIFT evaluation or statistical refitting. For an individual
sample, its GIFT calls and denominators can be inspected through the
[multi-sample dataset workflow](multi-sample-datasets.html).
