---
title: "When a genome is incomplete"
description: >
  Gene-loss experiments show how complete GIFT calls, missing-reaction
  diagnostics and assessable denominators behave on fragmented genomes.
---

```{r setup, include = FALSE}
# This case study reads committed R7 tables. It does not subsample genomes,
# re-evaluate GIFTs or refit the context model while the site is built.
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  dev = "ragg_png", dpi = 150, fig.retina = 2, out.width = "100%",
  fig.width = 9, fig.height = 5.5, fig.align = "center"
)
default_plot_hook <- knitr::knit_hooks$get("plot")
knitr::knit_hooks$set(plot = function(x, options) {
  href <- paste0("incomplete-genomes_files/figure-html/", basename(x[[1L]]))
  paste0(
    '<a class="case-figure" href="', href, '" target="_blank" ',
    'title="Open at full resolution">', default_plot_hook(x, options), "</a>"
  )
})

library(ggplot2)
root <- normalizePath(".")
while (!file.exists(file.path(root, "DESCRIPTION"))) root <- dirname(root)
out <- file.path(root, "manuscript", "analysis", "output")
read_tsv <- function(file) {
  utils::read.delim(file.path(out, file), check.names = FALSE,
                    stringsAsFactors = FALSE, quote = "")
}
decay <- read_tsv("incompleteness-decay.tsv")
summary <- read_tsv("incompleteness-summary.tsv")
assess <- read_tsv("incompleteness-assessability.tsv")
contexts <- read_tsv("cooccurrence-summary.tsv")
blocks <- read_tsv("block-drop-summary.tsv")

lost_n <- sum(decay$lost)
one_short_n <- sum(decay$one_reaction_short)
below <- assess[assess$policy == "completeness" & assess$kept < 0.9, ]
below_fraction <- below[below$metric == "supported_fraction", ]
below_assessable <- below[below$metric == "assessable_fraction", ]
zero_denominator_n <- sum(below_assessable$numerator == 0)
stopifnot(
  length(unique(decay$org)) == 120L,
  lost_n == 42636L,
  nrow(below_fraction) == 3823L,
  all(below_fraction$value == 1),
  zero_denominator_n == 497L,
  sum(contexts$n) == 718L,
  blocks$survivors_measured[blocks$block_mean == 20] == 2L
)
at_level <- function(kept, column) summary[[column]][summary$kept == kept]
fmt <- function(x, digits = 3) format(round(x, digits), nsmall = digits, trim = TRUE)
big <- function(x) format(x, big.mark = ",", trim = TRUE)

ink <- "#173330"; muted <- "#687672"; line <- "#dcded5"; surface <- "#fffdf8"
forest <- "#0d3d38"; coral <- "#ef765d"; blue <- "#477e91"
theme_case <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      text = element_text(colour = ink),
      axis.text = element_text(colour = muted),
      axis.title = element_text(colour = ink),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = line, linewidth = 0.35),
      plot.background = element_rect(fill = surface, colour = NA),
      panel.background = element_rect(fill = surface, colour = NA),
      plot.title = element_text(face = "bold"),
      plot.margin = margin(10, 14, 8, 10)
    )
}
```

<p class="lead">What does an incomplete genome do to a GIFT call? We removed
genes from complete reference genomes, then evaluated the remaining evidence
at each level. Complete calls disappear as their required steps lose support.
The closest incomplete route still names what is missing, while genome quality
changes only whether an unsupported call can count as an informative absence.</p>

<div class="case-stats">
<div><strong>`r length(unique(decay$org))`</strong><span>reference genomes</span></div>
<div><strong>`r big(lost_n)`</strong><span>calls lost in gene subsamples</span></div>
<div><strong>`r fmt(100 * one_short_n / lost_n, 1)`%</strong><span>lost calls one reaction short</span></div>
<div><strong>`r big(zero_denominator_n)`</strong><span>cases with no assessable frame member</span></div>
</div>

These experiments used database **2026.27.1**. The live catalogue has since
changed; the numbers below describe the pinned analysis, not a prediction for
the current database.

::: {.case-summary}
**In short**

- A complete route call falls faster under gene loss than a flat marker
  fraction, because losing one required reaction loses the call.
- Most lost calls are exactly one reaction short. They remain **unsupported**,
  but identify the chemistry to check in the assembly or annotation.
- The completeness policy withholds uninformative absences from denominators.
  It never restores a lost call. When no member is assessable, it emits no
  `supported_fraction`.
- Co-occurrence of neighboring genes did not justify filling missing steps.
:::

## What disappears when genes disappear

The analysis sampled 120 KEGG reference genomes across the curated-marker-count
distribution. For each genome, it retained 90%, 80%, 70%, 60% or 50% of its
genes in three independent subsamples, then evaluated the GIFTs again. Genes
were removed, and their markers followed. This preserves the fact that several
markers on one gene vanish together.

The comparator is the fraction of each GIFT's curated KO markers still present
in the same subsamples. It is a **flat marker-fraction baseline**, not native
KEGG MODULE evaluation, which also represents alternatives and complexes.

```{r retention, fig.cap = "Mean retained complete GIFT calls and mean flat marker-fraction score, each relative to its full-genome value on the same 120 genomes. The flat score is not KEGG MODULE completeness."}
retention <- rbind(
  data.frame(kept = summary$kept, value = summary$retention,
             measure = "Complete GIFT calls"),
  data.frame(kept = summary$kept, value = summary$fullness_ratio,
             measure = "Flat marker fraction")
)
retention$measure <- factor(retention$measure,
                            levels = c("Complete GIFT calls", "Flat marker fraction"))
ggplot(retention, aes(kept, value, colour = measure)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  scale_colour_manual(values = c("Complete GIFT calls" = forest,
                                 "Flat marker fraction" = coral)) +
  scale_x_continuous(breaks = summary$kept,
                     labels = paste0(100 * summary$kept, "%")) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  labs(x = "Genes retained", y = "Fraction of full-genome value", colour = NULL) +
  theme_case()
```

At 90% gene content, `r fmt(at_level(0.9, "retention"))` of the original complete
calls remain, while the marker-fraction score retains
`r fmt(at_level(0.9, "fullness_ratio"))` of its full-genome value. At 50%, the
values are `r fmt(at_level(0.5, "retention"))` and
`r fmt(at_level(0.5, "fullness_ratio"))`. A call is lost when a required
reaction loses its final accepted marker; a fraction can decline gradually.
No subsample gained a GIFT that the full gene set did not support.

## A lost call still names its gap

Across the subsamples, `r big(lost_n)` formerly complete calls were lost.
`r big(one_short_n)` of them, or `r fmt(100 * one_short_n / lost_n, 1)`%, lacked
exactly one reaction in their closest route. At 90% gene content, the share was
`r fmt(100 * at_level(0.9, "one_reaction_short") /
          at_level(0.9, "lost_calls"), 1)`%.

An unsupported call reports `minimum_missing_requirements` and its closest
implementation. For a metabolic GIFT, `missing_reactions_best_route` names the
reaction to investigate. The [single-genome guide](evaluating-a-genome.html#step-4-read-the-incomplete-calls-because-they-are-informative)
shows how to read that trace. A near miss is a curation or annotation lead; it
does not establish that the organism encodes the missing step.

## Quality changes a denominator, not a call

The same subsamples were read within three bounded anabolic frames at a
`threshold = 0.9`. Below the threshold, `policy = "completeness"` treats every
unsupported member as indeterminate for denominator calculations. In the
`r big(nrow(below_fraction))` below-threshold cases where
`supported_fraction` could still be reported, it was 1.000 every time: only
supported members remained assessable. In another
`r big(zero_denominator_n)` cases, no member was assessable and the
`supported_fraction` row was **absent**, rather than zero or `NA`.

Read `supported_fraction` alongside `assessable_fraction` and the supported
count. The latter never changes under the policy. The
[quantitative-traits tutorial](quantitative-traits.html#step-6-mag-completeness-and-the-honesty-of-a-denominator)
shows how to request and interpret those metrics for a MAG.

## Why neighboring evidence does not fill a gap

A second analysis asked whether a missing component or requirement could be
inferred from the others commonly found beside it. In the reference panel,
`r sum(contexts$n)` contexts had enough genomes for that screen. After controls
for genus and background prevalence, 43 appeared strong enough under an
independent-gene-loss assumption. Removing contiguous runs of neighboring
genes, closer to how assemblies lose contigs, left
`r blocks$survivors_measured[blocks$block_mean == 20]` candidates at a mean run
of 20 genes and
`r blocks$survivors_measured[blocks$block_mean == 50]` at 50 genes; those
candidates were not the same. Co-occurrence therefore provides no reliable rule
for changing an unsupported GIFT to supported.

Random gene deletion does not reproduce all assembly biases, and the experiment
kept the original gene calls fixed. Its retention curve describes a controlled
sensitivity analysis, not a calibration of real MAG completeness. The
contiguous-run analysis addresses gene linkage for the gap-filling question,
not every bias in an assembled genome.

## Reproduce this analysis

The figures and values here read the committed
[R7 tables](https://github.com/alberdilab/gifter/tree/main/manuscript/analysis/output).
The producing scripts are
[`02-incompleteness.R`](https://github.com/alberdilab/gifter/blob/main/manuscript/analysis/02-incompleteness.R),
[`09-cooccurrence.R`](https://github.com/alberdilab/gifter/blob/main/manuscript/analysis/09-cooccurrence.R)
and [`10-block-drop.R`](https://github.com/alberdilab/gifter/blob/main/manuscript/analysis/10-block-drop.R).
The [manuscript account](https://github.com/alberdilab/gifter/blob/main/manuscript/manuscript.md#r7-behaviour-under-genome-incompleteness)
gives the full methods, controls and interpretation.
