---
title: "Against observed phenotypes"
description: >
  Strain-matched BacDive observations, defined media and an independent trait
  synthesis show where GIFT calls agree with measured phenotypes and where
  their evidence runs out.
---

```{r setup, include = FALSE}
# This case study reads committed R9 results. It does not re-fetch phenotypes,
# annotate genomes or re-evaluate GIFTs while the documentation is built.
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  dev = "ragg_png", dpi = 150, fig.retina = 2, out.width = "100%",
  fig.width = 9, fig.height = 6, fig.align = "center"
)
default_plot_hook <- knitr::knit_hooks$get("plot")
knitr::knit_hooks$set(plot = function(x, options) {
  href <- paste0("phenotype-validation_files/figure-html/", basename(x[[1L]]))
  paste0(
    '<a class="case-figure" href="', href, '" target="_blank" ',
    'title="Open at full resolution">', default_plot_hook(x, options), "</a>"
  )
})

library(ggplot2)

root <- normalizePath(".")
while (!file.exists(file.path(root, "DESCRIPTION"))) root <- dirname(root)
out <- file.path(root, "manuscript", "analysis", "output")
read_tsv <- function(file) {
  utils::read.delim(file.path(out, file), check.names = FALSE,
                    stringsAsFactors = FALSE, quote = "")
}
fmt <- function(x, digits = 3) format(round(x, digits), nsmall = digits, trim = TRUE)
big <- function(x) format(x, big.mark = ",", trim = TRUE)

agreement <- read_tsv("phenotype-agreement.tsv")
failures <- read_tsv("phenotype-disagreements.tsv")
consistency <- read_tsv("phenotype-reference-consistency.tsv")
auxotrophy <- read_tsv("auxotrophy-agreement.tsv")
madin <- read_tsv("madin-agreement.tsv")
annotation <- read_tsv("annotation-route-calls.tsv")
coverage <- read_tsv("phenotype-validation-summary.tsv")
coverage_detail <- read_tsv("phenotype-validation-coverage.tsv")
manifest <- read_tsv("phenotype-validation-input-manifest.tsv")
release <- unique(manifest$database_version[manifest$input_kind == "gifter_database"])
stopifnot(length(release) == 1L, nzchar(release))

primary <- agreement[agreement$recall_usable & agreement$n >= 20, ]
status_n <- function(status) sum(coverage$gift_n[coverage$validation_status == status])
catalogue_n <- sum(coverage$gift_n)
individual_n <- status_n("individual_recall_usable")
frame_n <- status_n("bounded_frame_aggregate")
other_n <- catalogue_n - individual_n - frame_n
frame_only_ids <- coverage_detail$gift_id[coverage_detail$bounded_frame_aggregate]
gift_ids_for <- function(classes) {
  rows <- auxotrophy$gift_id[auxotrophy$substrate_class %in% classes]
  unique(unlist(strsplit(rows, "|", fixed = TRUE)))
}
essential_ids <- gift_ids_for(c("amino_acid", "nucleotide"))
context_ids <- gift_ids_for(c("cofactor", "inorganic_nitrogen"))
unmeasured_ids <- setdiff(frame_only_ids, c(essential_ids, context_ids))
stopifnot(length(frame_only_ids) == length(essential_ids) +
            length(context_ids) + length(unmeasured_ids),
          setequal(frame_only_ids, c(essential_ids, context_ids, unmeasured_ids)),
          identical(unmeasured_ids, "siroheme_to_heme_b"))
pick <- function(table, term, field = NULL, kinds = NULL) {
  x <- table[table$term == term, ]
  if (!is.null(field)) x <- x[x$field == field, ]
  if (!is.null(kinds)) x <- x[grepl(kinds, x$kinds), ]
  stopifnot(nrow(x) == 1L)
  x
}

ink <- "#173330"; muted <- "#687672"; line <- "#dcded5"; surface <- "#fffdf8"
blue <- "#477e91"
theme_case <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      text = element_text(colour = ink),
      axis.text = element_text(colour = muted),
      axis.title = element_text(colour = ink),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = line, linewidth = 0.35),
      plot.background = element_rect(fill = surface, colour = NA),
      panel.background = element_rect(fill = surface, colour = NA),
      strip.text = element_text(colour = ink, face = "bold", hjust = 0),
      plot.title = element_text(face = "bold"),
      plot.margin = margin(10, 14, 8, 10)
    )
}
```

<p class="lead">When a strain is observed to use a substrate, does its genome
support the corresponding GIFT? We compared curated genomic calls with
strain-resolved BacDive assays, growth on chemically defined MediaDive media,
and a second, species-level trait compilation. The result is a test of where
the <em>genomic evidence</em> reaches an observed capability, not a prediction
of what a strain will do in any particular assay.</p>

<div class="case-stats">
<div><strong>`r catalogue_n`</strong><span>GIFTs in the tested release</span></div>
<div><strong>`r nrow(primary)`</strong><span>BacDive comparisons with at least 20 paired records</span></div>
<div><strong>`r individual_n`</strong><span>GIFTs individually testable at that threshold</span></div>
<div><strong>`r big(sum(auxotrophy$n[auxotrophy$substrate_class %in% c("amino_acid", "nucleotide")]))`</strong><span>defined-medium tests on essential nutrient classes</span></div>
</div>

These results are pinned to database **`r release`** and its committed analysis
tables. The live catalogue can contain additional GIFTs or changed routes.

::: {.case-summary}
**In short**

- Recall is reported separately for each observed phenotype. An EC-resolved
  enzyme assay tests a reaction; substrate use and product formation test a GIFT.
- A positive genomic call with no observed phenotype is allowed: expression,
  regulation and assay conditions lie outside the GIFT claim. An observed
  capability with no supported call is the useful disagreement to investigate.
- Only `r individual_n` of `r catalogue_n` GIFTs have a direct individual
  recall test at this threshold. Another `r frame_n` belong to bounded
  anabolic frames: defined-medium growth can challenge their nutrient-level
  support, with limits explained below. The remaining `r other_n` have no
  usable test in these references.
:::

## What was compared

BacDive records measured strain phenotypes and assembly accessions. We joined
those accessions to the reference genomes, mapped metabolite identifiers to
declared anchors and EC identifiers to reactions, and excluded BacDive's
**genome-based predictions**. The reviewed crosswalk admits a phenotype to a
recall calculation only where its boundary is `equivalent` to, or narrower
than, the GIFT claim. For example, gelatin hydrolysis is not treated as an
observation of collagen cleavage, and motility is not treated as an observation
of a flagellar apparatus.

For an observation that passes that rule, recall is the share of
**observed-positive** records whose matched genomes support the target. The
`n` beside each point below counts *all* paired observations, positive and
negative; the recall denominator is the observed-positive subset. Each row
also gives the number of represented genera. Rows are kept separate even when
they name the same substrate, because growth on a sugar and acid production
from it are different assays.

```{r primary-recall, fig.width = 9.5, fig.height = 8.5, fig.cap = "BacDive recall by curated observation, with all paired records (n) and represented genera shown at right. Only equivalent or subset-of mappings with at least 20 pairs appear. The positive subset of each n is the recall denominator. No rate is pooled across observations."}
primary$assay <- ifelse(
  primary$field == "enzymes", "enzyme activity",
  ifelse(primary$field == "metabolite production", "product formation",
         ifelse(grepl("builds acid", primary$kinds), "acidification", "substrate use"))
)
primary$label <- paste(primary$term, primary$assay, sep = " · ")
primary$layer_label <- ifelse(primary$layer == "reaction",
                              "Reaction: enzyme activity",
                              "GIFT: substrate use or product formation")
primary <- primary[order(primary$layer, primary$recall), ]
primary$label <- factor(primary$label, levels = unique(primary$label))

ggplot(primary, aes(recall, label)) +
  geom_segment(aes(x = 0, xend = recall, yend = label),
               colour = line, linewidth = 0.8) +
  geom_point(colour = blue, size = 2.2) +
  geom_text(aes(x = 1.03, label = sprintf("n = %d · %d genera", n, genera)),
            hjust = 0, size = 2.5, colour = muted) +
  facet_wrap(~ layer_label, ncol = 1, scales = "free_y") +
  scale_x_continuous(limits = c(0, 1.62), breaks = seq(0, 1, 0.25)) +
  labs(x = "recall among observed-positive records", y = NULL) +
  theme_case() +
  theme(panel.grid.major.y = element_blank(),
        axis.text.y = element_text(size = 9))
```

Urease activity reaches `r fmt(pick(primary, "urease", "enzymes")$recall)`
recall across `r big(pick(primary, "urease", "enzymes")$n)` paired records;
urea utilisation reaches
`r fmt(pick(primary, "urea", "metabolite utilization")$recall)` across
`r big(pick(primary, "urea", "metabolite utilization")$n)`. Indole production
reaches `r fmt(pick(primary, "indole")$recall)`. N-acetylglucosamine use
reaches only `r fmt(pick(primary, "N-acetylglucosamine")$recall)`: tracing
its missing reaction led to curation of a second route through the
phosphotransferase system. That is a specific route lead, not evidence that
the phenotype itself was inferred from the genome.

## A disagreement has a direction

The `r big(sum(primary$encoded_not_observed))` supported-call/negative-assay
comparisons are **permitted** under the GIFT claim. Of the
`r big(sum(primary$observed_not_encoded))` observed-positive/unsupported-call
comparisons, `r sum(failures$class[failures$target %in% primary$target] == "incomplete route")`
have a closest route with a named missing requirement and
`r sum(failures$class[failures$target %in% primary$target] == "no evidence")`
have no genomic evidence for that capability. These are counts of
genome–observation comparisons, not distinct genomes. A missing requirement
is a curation lead; it is not automatically an annotation error or a missing
enzyme.

Two assays in BacDive sometimes disagree with each other. For urease activity
versus urea utilisation, `r consistency$agree_with_genome[consistency$gift_id == "urea_hydrolysis"]`
of `r consistency$strains_with_genome[consistency$gift_id == "urea_hydrolysis"]`
genome-backed strains agree. For tryptophanase activity versus indole
production, only
`r consistency$agree_with_genome[consistency$gift_id == "tryptophan_degradation_indole"]`
of `r consistency$strains_with_genome[consistency$gift_id == "tryptophan_degradation_indole"]`
agree. An unexpectedly low recall on one assay needs to be read beside that
assay's own consistency with the reference.

## A second reference checks the mapping

Madin's published species-level compilation records carbon-substrate use. It
offers an independent check on whether BacDive's *growth* and *acidification*
rows should be kept separate. The comparison below uses the reported recall
for each source; the strain and species denominators differ, so the columns
are not paired measurements of the same organisms.

```{r substrate-comparison}
sugar_targets <- data.frame(
  substrate = c("L-arabinose", "D-galactose", "L-rhamnose", "D-xylose"),
  madin_term = c("arabinose", "galactose", "rhamnose", "xylose")
)
sugar_targets$bacdive_growth <- vapply(sugar_targets$substrate, function(term)
  pick(agreement, term, "metabolite utilization", "carbon source")$recall,
  numeric(1))
sugar_targets$bacdive_acid <- vapply(sugar_targets$substrate, function(term)
  pick(agreement, term, "metabolite utilization", "builds acid")$recall,
  numeric(1))
sugar_targets$madin_use <- vapply(sugar_targets$madin_term, function(term)
  pick(madin, term, "carbon_substrates")$recall,
  numeric(1))
knitr::kable(
  data.frame(
    Substrate = sugar_targets$substrate,
    `BacDive growth/use` = fmt(sugar_targets$bacdive_growth),
    `BacDive acidification` = fmt(sugar_targets$bacdive_acid),
    `Madin substrate use` = fmt(sugar_targets$madin_use),
    check.names = FALSE
  ),
  caption = "Recall from distinct observed-positive sets; each BacDive test kind remains a separate claim."
)
```

The independent substrate-use values track BacDive growth/use more closely
than acidification. Madin requires a species-to-genome representative choice;
the [analysis](https://github.com/alberdilab/gifter/blob/main/manuscript/analysis/07-madin.R)
repeated that choice 100 times and recorded its sensitivity. Its motility row,
like BacDive's, is excluded from flagellar-apparatus recall because motility
does not imply a flagellum.

## What defined media and annotation add

MediaDive connects growth-positive strains to chemically defined media. If a
medium omits a biomass-essential nutrient and the strain grows, that is an
observed-positive test of the relevant biosynthetic capability. Alternative
curated routes to the same nutrient are OR-ed: growth without cysteine asks
whether the genome supports **either** curated cysteine route, not whether it
supports both. Only media whose ingredients are at least 75% resolvable to
identifiers enter this inference. For each growth-positive strain–medium pair,
the analysis records which omitted nutrients have genomic support; it then
summarises support by nutrient and by genome within a bounded frame. An
unsupported product points to GIFT evidence worth tracing. A medium that
supplies a nutrient gives no test of its synthesis.

This gives the strongest indirect check for `r length(essential_ids)` of the
`r frame_n` frame-only GIFTs: they make `r sum(auxotrophy$substrate_class %in% c("amino_acid", "nucleotide"))`
amino-acid or nucleotide products. Across those products, the committed table
records
`r big(sum(auxotrophy$n[auxotrophy$substrate_class %in% c("amino_acid", "nucleotide")]))`
nutrient-level tests and
`r fmt(sum(auxotrophy$supported[auxotrophy$substrate_class %in% c("amino_acid", "nucleotide")]) / sum(auxotrophy$n[auxotrophy$substrate_class %in% c("amino_acid", "nucleotide")]))`
recall. The [per-nutrient table](https://github.com/alberdilab/gifter/blob/main/manuscript/analysis/output/auxotrophy-agreement.tsv)
shows where support is strong or weak; the [disagreement table](https://github.com/alberdilab/gifter/blob/main/manuscript/analysis/output/auxotrophy-disagreements.tsv)
retains the genomes, media and omitted nutrients behind unsupported calls.

The other `r length(context_ids)` measured frame-only GIFTs make cofactors or
inorganic nitrogen products. Their nutrient-level rows are reported, but a
strain may grow without making a particular cofactor or without fixing
nitrogen, so a low support fraction is not by itself a failed genomic call.
The final frame member, `siroheme_to_heme_b`, was excluded from this test
because the KO-only genome annotation cannot evidence one of its components.
Thus **`r frame_n` counts frame membership**, not 44 independent phenotype
assays or 44 equally falsifiable GIFTs. Even where a nutrient test is sound,
growth implies some way to obtain that product; it does not identify which
curated route the organism uses.

The primary comparison uses KEGG's own KO assignments. On the same
`r unique(annotation$genomes)` genomes, a pinned annotation pipeline recovered
`r fmt(sum(annotation$annot_ko) / sum(annotation$kegg_ko))` of the complete
KEGG-route calls using KofamScan alone, and
`r fmt(sum(annotation$annot_full) / sum(annotation$kegg_ko))` after adding
NCBIfam, dbCAN and Pfam evidence. Those fractions describe the change in
**total supported calls**, not per-GIFT recall or overall accuracy; some
capabilities gained calls while others lost them. Gene calling was held fixed,
so this comparison isolates marker assignment rather than all losses a MAG
annotation workflow could introduce.

## How much of the catalogue could be tested?

For this `r catalogue_n`-GIFT release, `r individual_n` GIFTs had a direct
individual recall test at 20 or more paired observations. Another `r frame_n`
had no such assay comparison but belonged to bounded anabolic frames;
`r length(frame_only_ids) - length(unmeasured_ids)` entered the indirect
defined-medium analysis described above. The other `r other_n` had no usable
recall test in these references: some had no public observation, while others
had too few records or only a related or refused proxy. This is a limit of the
available observations and crosswalk, not a negative GIFT call.

The [coverage table](https://github.com/alberdilab/gifter/blob/main/manuscript/analysis/output/phenotype-validation-coverage.tsv)
names the status of every GIFT. The [full manuscript account](https://github.com/alberdilab/gifter/blob/main/manuscript/manuscript.md#r9-agreement-with-observed-phenotypes)
sets out the comparison and its limitations.

## Reproduce this analysis

All figures and values on this page read the committed
[R9 tables](https://github.com/alberdilab/gifter/tree/main/manuscript/analysis/output).
The producing scripts are
[`03-phenotype.R`](https://github.com/alberdilab/gifter/blob/main/manuscript/analysis/03-phenotype.R),
[`04-auxotrophy.R`](https://github.com/alberdilab/gifter/blob/main/manuscript/analysis/04-auxotrophy.R),
[`05-annotation-route.R`](https://github.com/alberdilab/gifter/blob/main/manuscript/analysis/05-annotation-route.R),
[`07-madin.R`](https://github.com/alberdilab/gifter/blob/main/manuscript/analysis/07-madin.R)
and the [coverage audit](https://github.com/alberdilab/gifter/blob/main/manuscript/analysis/19-phenotype-validation-coverage.R).
The reviewed [phenotype crosswalk](https://github.com/alberdilab/gifter/blob/main/data-raw/reference/phenotype-crosswalk.tsv)
states exactly which observation licenses each comparison.
