---
title: "Across the bacterial phylogeny"
description: >
  Encoded GIFT repertoires of 696 complete genomes spanning every eligible
  order of the GTDB bacterial tree.
---

```{r setup, include = FALSE}
# A case study reports a finished analysis. Nothing is evaluated here: every
# number and figure is read from the tables that manuscript/analysis/ committed,
# so the page cannot drift from the scripts that produced them.
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  dev = "ragg_png", dpi = 150, fig.retina = 2, out.width = "100%",
  fig.width = 9, fig.height = 5.4, fig.align = "center"
)

# Each figure links to its own full-resolution file.
default_plot_hook <- knitr::knit_hooks$get("plot")
knitr::knit_hooks$set(plot = function(x, options) {
  href <- paste0("phylogeny_files/figure-html/", basename(x[[1L]]))
  paste0(
    '<a class="case-figure" href="', href, '" target="_blank" ',
    'title="Open at full resolution">', default_plot_hook(x, options), "</a>"
  )
})

library(ggplot2)
library(patchwork)

root <- normalizePath(".")
while (!file.exists(file.path(root, "DESCRIPTION"))) root <- dirname(root)
output_dir <- file.path(root, "manuscript", "analysis", "output")
panel_dir <- file.path(root, "manuscript", "analysis", "gtdb-phylogeny")

read_tsv <- function(dir, file) {
  utils::read.delim(
    file.path(dir, file), check.names = FALSE, stringsAsFactors = FALSE,
    quote = ""
  )
}
item <- function(x, name) x$value[match(name, x$item)]
strip_rank <- function(x) sub("^[a-z]__", "", x)
big <- function(x) format(as.numeric(x), big.mark = ",", trim = TRUE)

audit <- read_tsv(output_dir, "gtdb-phylogeny-input-audit.tsv")
selection <- read_tsv(panel_dir, "selection-audit.tsv")
origins <- read_tsv(panel_dir, "origin-group-summary.tsv")
genomes <- read_tsv(output_dir, "gtdb-phylogeny-genome-summary.tsv")
gifts <- read_tsv(output_dir, "gtdb-phylogeny-gift-summary.tsv")
phyla <- read_tsv(output_dir, "gtdb-phylogeny-phylum-summary.tsv")
expectation <- read_tsv(output_dir, "gtdb-size-expectation-genomes.tsv")
model <- read_tsv(output_dir, "gtdb-size-expectation-model.tsv")
taxa <- read_tsv(output_dir, "gtdb-size-expectation-taxa.tsv")
signal <- read_tsv(output_dir, "gtdb-size-deviation-phylogenetic-signal.tsv")
classes <- read_tsv(output_dir, "gtdb-gift-class-genome-size.tsv")
gift_signal <- read_tsv(output_dir, "gtdb-gift-phylogenetic-signal.tsv")
origin_traits <- read_tsv(output_dir, "gtdb-origin-genome-traits.tsv")
near_steps <- read_tsv(output_dir, "gtdb-near-miss-steps.tsv")
near_readings <- read_tsv(output_dir, "gtdb-near-miss-readings.tsv")
near_genomes <- read_tsv(output_dir, "gtdb-near-miss-genomes.tsv")
calls <- utils::read.delim(
  xzfile(file.path(output_dir, "gtdb-phylogeny-calls.tsv.xz")),
  check.names = FALSE, stringsAsFactors = FALSE, quote = ""
)
tree <- ape::read.tree(file.path(panel_dir, "selected-genomes.tree"))

n_genomes <- nrow(genomes)
n_gifts <- nrow(gifts)
supported <- genomes[["all supported GIFTs"]]

# The site palette (pkgdown/extra.css).
ink <- "#173330"; muted <- "#687672"; line <- "#dcded5"; surface <- "#fffdf8"
forest <- "#0d3d38"; mint <- "#90c9b5"; coral <- "#ef765d"; gold <- "#d9a936"
blue <- "#477e91"

theme_case <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      text = element_text(colour = ink),
      axis.text = element_text(colour = muted),
      axis.title = element_text(colour = ink, size = rel(0.92)),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = line, linewidth = 0.35),
      plot.background = element_rect(fill = surface, colour = NA),
      panel.background = element_rect(fill = surface, colour = NA),
      legend.background = element_rect(fill = surface, colour = NA),
      legend.key = element_rect(fill = surface, colour = NA),
      legend.position = "bottom",
      legend.title = element_text(size = rel(0.9)),
      strip.text = element_text(colour = ink, face = "bold", hjust = 0),
      plot.title = element_text(face = "bold", size = rel(1)),
      plot.margin = margin(10, 14, 8, 10)
    )
}

# One colour per major phylum, shared by every figure on the page.
phylum_counts <- sort(table(strip_rank(genomes$phylum)), decreasing = TRUE)
major_phyla <- names(phylum_counts)[1:12]
phylum_levels <- c(major_phyla, "other phyla")
phylum_colours <- setNames(
  c(grDevices::hcl.colors(length(major_phyla), "Dark 3"), "#cfcec6"),
  phylum_levels
)
phylum_group <- function(x) {
  x <- strip_rank(x)
  factor(ifelse(x %in% major_phyla, x, "other phyla"), levels = phylum_levels)
}
```

<p class="lead">How much of what a bacterium encodes is set by the size of its
genome, and how much by its lineage? This case study evaluates the whole GIFT
catalogue on `r n_genomes` complete genomes that span every eligible order of
the GTDB bacterial tree.</p>

<div class="case-stats">
<div><strong>`r n_genomes`</strong><span>complete genomes</span></div>
<div><strong>`r item(audit, "selected phyla")`</strong><span>phyla</span></div>
<div><strong>`r item(audit, "selected orders")`</strong><span>orders</span></div>
<div><strong>`r n_gifts`</strong><span>GIFTs evaluated</span></div>
<div><strong>`r big(item(audit, "supported genome-GIFT pairs"))`</strong><span>supported calls</span></div>
</div>

::: {.case-summary}
**In short**

- Genome size alone explains `r round(100 * as.numeric(item(model, "deviance explained")))`%
  of the variation in how many GIFTs a genome supports.
- What remains is lineage: related genomes sit above or below the size
  expectation together (Pagel's λ = `r signif(signal$pagel_lambda[[1]], 2)`).
- Catabolic capabilities follow genome size more closely than biosynthetic
  ones, and are scattered more widely over the tree.
- Where a genome is one requirement short of a GIFT, the same requirement is
  usually the one missing, which gives us a ranked list of GIFT definitions
  to revisit.
:::

If you have not yet read what a GIFT call claims,
[start with a single genome](evaluating-a-genome.html).

## The genome panel

The panel was drawn from the `r big(item(selection, "complete-genome species representatives"))`
species representatives of GTDB `r item(audit, "GTDB release")` that are
complete assemblies and tips of the bac120 reference tree. A genome was eligible
when its isolation source could be assigned to at least one origin group, which
left `r big(item(selection, "origin-classified eligible representatives"))`
candidates. Selection then ran in two steps, without looking at any annotation
or GIFT call:

1. **One genome per order.** The phylogenetic medoid of each of the
   `r item(audit, "selected orders")` eligible orders, so that no order is
   missing.
2. **Balanced origins.** Further genomes were added until every origin group
   reached 85 genomes, or all of its members if it had fewer, each time taking
   the genome that added the most phylogenetic diversity.

```{r origin-table}
origin_table <- origins[origins$origin_group != "any_origin_group", ]
knitr::kable(
  data.frame(
    Origin = origin_table$description,
    Eligible = big(origin_table$eligible_genomes),
    Selected = origin_table$selected_genomes
  ),
  align = "lrr",
  caption = "Origin groups of the panel. A genome can belong to several, so the counts do not add up to the panel size."
)
```

The panel is deliberately enriched for breadth. Its numbers describe these
`r n_genomes` genomes and are not prevalence estimates for bacteria at large.

## From annotation to calls

Each genome was annotated with [giftag](https://alberdilab.github.io/giftag/)'s
marker sources through Drakkar, which produced
`r sprintf("%.1f", as.numeric(item(audit, "marker rows")) / 1e6)` million marker
rows. The whole panel was then evaluated in one call, as a community of
`r n_genomes` genomes:

```{r, eval = FALSE, echo = TRUE}
library(gifter)

# markers: one row per observed marker, with genome_id, gene_id, namespace
# and accession
calls <- evaluate_gifts_community(markers)
```

The result holds one row per genome and GIFT, `r big(item(audit, "genome-GIFT calls"))`
in total, each with its closest implementation, the requirements it lacks and
the genes behind it. Everything below is read from that table
(database `r item(audit, "database version")`, gifter `r item(audit, "package version")`).

## Repertoires across the tree

```{r overview-data}
# Tip order and coordinates of a rectangular tree, from the edge matrix.
tree <- ape::reorder.phylo(tree, "cladewise")
n_tips <- length(tree$tip.label)
node_x <- ape::node.depth.edgelength(tree)
node_y <- numeric(n_tips + tree$Nnode)
tip_edges <- tree$edge[tree$edge[, 2] <= n_tips, 2]
node_y[tip_edges] <- seq_len(n_tips)
post <- ape::reorder.phylo(tree, "postorder")$edge
for (parent in unique(post[, 1])) {
  node_y[parent] <- mean(range(node_y[post[post[, 1] == parent, 2]]))
}
tree_h <- data.frame(
  x = node_x[tree$edge[, 1]], xend = node_x[tree$edge[, 2]],
  y = node_y[tree$edge[, 2]]
)
tree_v <- do.call(rbind, lapply(unique(tree$edge[, 1]), function(parent) {
  ys <- node_y[tree$edge[tree$edge[, 1] == parent, 2]]
  data.frame(x = node_x[parent], y = min(ys), yend = max(ys))
}))
tips <- data.frame(genome_id = tree$tip.label, y = node_y[seq_len(n_tips)])
tips <- merge(tips, genomes[c("genome_id", "phylum", "all supported GIFTs")])
tips$phylum_group <- phylum_group(tips$phylum)
tips <- tips[order(tips$y), ]

runs <- rle(as.character(tips$phylum_group))
run_end <- cumsum(runs$lengths)
runs <- data.frame(
  phylum_group = runs$values, n = runs$lengths,
  y = (run_end - runs$lengths + 1 + run_end) / 2
)
runs <- runs[runs$phylum_group != "other phyla", ]
phylum_labels <- do.call(rbind, lapply(
  split(runs, runs$phylum_group), function(r) r[which.max(r$n), ]
))

heat <- merge(calls[c("genome_id", "gift_id", "gift_type", "complete",
                      "evidence_confidence")], tips[c("genome_id", "y")])
heat$gift_id <- factor(heat$gift_id, levels = gifts$gift_id[order(gifts$heatmap_column)])
heat$gift_type <- factor(
  heat$gift_type, levels = c("metabolic", "structural", "regulatory", "defense")
)
heat$status <- ifelse(heat$complete, heat$evidence_confidence, "unsupported")
heat$status <- factor(
  heat$status, levels = c("unsupported", "ambiguous", "high-confidence", "curated")
)
```

```{r overview, fig.width = 10, fig.height = 11.5, fig.cap = "Every GIFT call of the panel. Rows are genomes in the order of the pruned GTDB tree, with the twelve largest phyla named; columns are GIFTs, grouped by type (metabolic, structural, regulatory, defense). A filled cell is a genome that supports a complete implementation, coloured by the confidence of its evidence. The bars count the supported GIFTs of each genome."}
shared_y <- scale_y_continuous(limits = c(0.5, n_tips + 0.5), expand = c(0, 0))
void <- theme_void(base_size = 12) +
  theme(
    plot.background = element_rect(fill = surface, colour = NA),
    text = element_text(colour = ink),
    plot.title = element_text(face = "bold", size = 11, hjust = 0,
                              margin = margin(b = 6))
  )

p_tree <- ggplot() +
  geom_segment(data = tree_h, aes(x = x, xend = xend, y = y, yend = y),
               colour = muted, linewidth = 0.2) +
  geom_segment(data = tree_v, aes(x = x, xend = x, y = y, yend = yend),
               colour = muted, linewidth = 0.2) +
  shared_y + labs(title = "GTDB tree") + void

p_names <- ggplot(phylum_labels, aes(x = 1, y = y, label = phylum_group)) +
  geom_text(hjust = 1, size = 3.1, colour = ink) +
  shared_y + scale_x_continuous(limits = c(0, 1), expand = c(0, 0.03)) +
  coord_cartesian(clip = "off") + labs(title = "") + void

p_strip <- ggplot(tips, aes(x = 1, y = y, fill = phylum_group)) +
  geom_raster() + shared_y +
  scale_fill_manual(values = phylum_colours, guide = "none") +
  labs(title = "") + void

p_heat <- ggplot(heat, aes(x = gift_id, y = y, fill = status)) +
  geom_raster() +
  facet_grid(cols = vars(gift_type), scales = "free_x", space = "free_x",
             labeller = as_labeller(c(
               metabolic = "metabolic", structural = "str.",
               regulatory = "reg.", defense = "def."
             ))) +
  shared_y +
  scale_fill_manual(
    values = c(unsupported = "#efece3", ambiguous = gold,
               `high-confidence` = "#5fa893", curated = forest),
    drop = FALSE, name = NULL
  ) +
  labs(title = paste(n_gifts, "GIFTs")) + void +
  theme(
    strip.text = element_text(size = 8.5, face = "bold", margin = margin(b = 3)),
    strip.clip = "off",
    panel.spacing.x = grid::unit(2.4, "mm"),
    legend.position = "bottom",
    legend.key.size = grid::unit(4, "mm"),
    legend.text = element_text(size = 10)
  )

p_count <- ggplot(tips, aes(x = `all supported GIFTs`, y = y)) +
  geom_segment(aes(x = 0, xend = `all supported GIFTs`, yend = y),
               colour = forest, linewidth = 0.35) +
  shared_y +
  scale_x_continuous(expand = expansion(mult = c(0, 0.04))) +
  labs(title = "Supported", x = NULL) +
  theme_case(11) +
  theme(
    axis.text.y = element_blank(), axis.title.y = element_blank(),
    panel.grid.major.y = element_blank(),
    plot.title = element_text(size = 11, margin = margin(b = 6)),
    plot.margin = margin(0, 8, 0, 4)
  )

p_tree + p_names + p_strip + p_heat + p_count +
  plot_layout(widths = c(1.2, 0.72, 0.07, 4.4, 0.8), guides = "collect") &
  theme(legend.position = "bottom")
```

```{r overview-numbers}
pate <- phyla[phyla$phylum == "p__Patescibacteriota", ]
top_genome <- genomes[which.max(supported), ]
never <- sum(gifts$complete == 0)
```

A genome supports a median of `r median(supported)` of the `r n_gifts` GIFTs,
and the range runs from `r min(supported)` to `r max(supported)`
(*`r strip_rank(top_genome$species)`*). The catalogue is well exercised:
`r n_gifts - never` GIFTs are supported by at least one genome, and
`r round(100 * mean(calls$evidence_confidence[calls$complete] == "curated"))`%
of the supported calls carry curated evidence confidence.

The pale bands are the lineages with reduced genomes. The
`r pate$genomes` Patescibacteriota support a median of
`r pate$median_supported` GIFTs, and the Bacillota_I lineage, which holds the
mycoplasmas, a median of
`r phyla$median_supported[phyla$phylum == "p__Bacillota_I"]`. At the other end,
the largest repertoires belong to Pseudomonadota. Outside the reduced lineages
some metabolic GIFTs form nearly solid columns, while many others appear only
in patches of related genomes.

## Genome size sets the expectation

```{r size, fig.height = 5.6, fig.cap = "Supported GIFTs against the size of the evaluated assembly. Each point is a genome, coloured by phylum. The line is the expectation from a smooth model of the supported share of the catalogue given genome size."}
size <- expectation
size$phylum_group <- phylum_group(size$phylum)
size <- size[order(size$phylum_group, decreasing = TRUE), ]
curve <- size[order(size$genome_size), ]

ggplot(size, aes(x = genome_size / 1e6, y = supported)) +
  geom_point(aes(colour = phylum_group), size = 1.7, alpha = 0.85) +
  geom_line(data = curve, aes(y = expected), colour = ink, linewidth = 0.9) +
  scale_x_log10(breaks = c(0.25, 0.5, 1, 2, 4, 8, 16),
                labels = function(x) format(x, drop0trailing = TRUE, trim = TRUE)) +
  scale_colour_manual(values = phylum_colours, name = NULL) +
  labs(x = "genome size (Mb, log scale)", y = "supported GIFTs") +
  guides(colour = guide_legend(nrow = 3, override.aes = list(size = 3, alpha = 1))) +
  theme_case()
```

The number of supported GIFTs rises with genome size
(Spearman ρ = `r round(as.numeric(item(model, "Spearman rho, genome size and supported GIFTs")), 2)`).
The relationship is not a straight line. Below about 1 Mb genomes support
almost nothing in the catalogue, between 1 and 5 Mb the repertoire grows
steeply, and above that it levels off at around 65 GIFTs.

The fitted curve is the **size expectation**: the number of GIFTs a genome of
that size supports on average in this panel. A genome's **deviation** is the
difference between what it supports and that expectation. No single genome
departs from it significantly once the `r n_genomes` tests are corrected for,
but lineages do.

## Lineages depart from it together

```{r taxa, fig.height = 6, fig.cap = "Deviation from the size expectation in each phylum with at least five panel genomes (count in brackets). A phylum is coloured when its genomes sit consistently above or below the expectation at a false discovery rate of 0.05."}
phylum_taxa <- taxa[taxa$rank == "phylum", ]
dev <- expectation[expectation$phylum %in% phylum_taxa$taxon, ]
dev <- merge(dev, phylum_taxa[c("taxon", "genomes", "median_residual", "deviation")],
             by.x = "phylum", by.y = "taxon", suffixes = c("", "_taxon"))
dev$label <- paste0(strip_rank(dev$phylum), " (", dev$genomes, ")")
dev$label <- reorder(dev$label, dev$median_residual)
deviation_colours <- c(
  "more than expected" = coral, "within expectation" = "#b9b8b0",
  "fewer than expected" = blue
)

ggplot(dev, aes(x = residual, y = label)) +
  geom_vline(xintercept = 0, colour = ink, linewidth = 0.4) +
  geom_boxplot(aes(fill = deviation_taxon, colour = deviation_taxon),
               alpha = 0.35, outlier.shape = NA, width = 0.62, linewidth = 0.4) +
  geom_point(aes(colour = deviation_taxon), size = 0.9, alpha = 0.7,
             position = position_jitter(height = 0.16, width = 0, seed = 1)) +
  scale_fill_manual(values = deviation_colours, name = NULL,
                    aesthetics = c("fill", "colour")) +
  labs(x = "observed − expected GIFTs", y = NULL) +
  theme_case() +
  theme(panel.grid.major.y = element_blank(),
        axis.text.y = element_text(colour = ink))
```

```{r taxa-numbers}
n_tested <- nrow(phylum_taxa)
n_more <- sum(phylum_taxa$deviation == "more than expected")
n_fewer <- sum(phylum_taxa$deviation == "fewer than expected")
more_names <- strip_rank(phylum_taxa$taxon[phylum_taxa$deviation == "more than expected"])
```

Of the `r n_tested` phyla with at least five genomes, `r n_more` support more
GIFTs than their genome sizes predict (`r paste(more_names, collapse = ", ")`)
and `r n_fewer` support fewer. The deviation is strongly structured by
phylogeny: Pagel's λ is `r signif(signal$pagel_lambda[[1]], 2)` and Blomberg's K
is `r signif(signal$blomberg_k[[1]], 2)`, and genomes deviate in the same
direction as their relatives out to a patristic distance of about 1.1
substitutions per site.

A phylum below the expectation does not necessarily encode less. We curated
the catalogue mostly from the biochemistry of well-studied organisms, so a
lineage that uses routes we have not curated yet will read as short. The
[last section](#one-step-short) looks at where genomes fall just short.

## Which capabilities scale with size

```{r classes, fig.height = 7.2, fig.cap = "Correlation between genome size and the number of supported GIFTs in each class with at least five GIFTs (count in brackets), with 95% bootstrap intervals. Classes come from the GIFT type, the metabolic mode and two curated facets, and a GIFT can belong to several."}
cls <- classes[classes$grouping != "all GIFTs", ]
group_labels <- c(
  gift_type = "GIFT type", mode = "Mode",
  physiological_role = "Physiological role", substrate_class = "Substrate class"
)
cls$grouping <- factor(group_labels[cls$grouping], levels = group_labels)
cls$label <- paste0(gsub("_", " ", cls$class), " (", cls$gifts, ")")
cls$label <- reorder(cls$label, cls$spearman_rho)

ggplot(cls, aes(x = spearman_rho, y = label)) +
  geom_vline(xintercept = classes$spearman_rho[classes$grouping == "all GIFTs"],
             colour = coral, linetype = "22", linewidth = 0.5) +
  geom_segment(aes(x = rho_lower, xend = rho_upper, yend = label),
               colour = muted, linewidth = 0.6) +
  geom_point(colour = forest, size = 2.6) +
  facet_grid(rows = vars(grouping), scales = "free_y", space = "free_y",
             switch = "y") +
  scale_x_continuous(limits = c(0, 0.85), expand = c(0, 0)) +
  labs(x = "Spearman ρ with genome size", y = NULL) +
  theme_case() +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(colour = ink),
    strip.placement = "outside",
    strip.text.y.left = element_text(angle = 90, hjust = 0.5, colour = muted,
                                     face = "plain", size = 10),
    panel.spacing.y = grid::unit(5, "mm")
  )
```

```{r class-numbers}
rho <- function(class) sprintf("%.2f", classes$spearman_rho[classes$class == class])
```

The dashed line is the correlation for the whole catalogue, and mode is the
metabolic mode of a GIFT. Catabolic GIFTs
follow genome size more closely (ρ = `r rho("catabolic")`) than anabolic ones
(ρ = `r rho("anabolic")`): the classes that follow size most closely are those for acquiring
carbon and nitrogen, while biosynthetic classes are already widely supported in
mid-sized genomes. Two kinds of capability barely follow size at all, structural machinery
(ρ = `r rho("structural")`) and fermentative end products
(ρ = `r rho("fermentative_end_product")`), which are properties of a lineage's
way of life and not of how much it encodes.

## How capabilities are distributed over the tree

```{r signal, fig.height = 4.8, fig.cap = "Phylogenetic signal of each GIFT that at least ten genomes support and at least ten lack, as Fritz and Purvis's D. A GIFT near 0 is as clumped on the tree as a trait inherited by descent; a GIFT near 1 is scattered at random. Bars are class medians."}
sig <- gift_signal
sig$class <- ifelse(sig$gift_type == "metabolic",
                    paste0("metabolic, ", sig$mode), sig$gift_type)
medians <- stats::aggregate(d_statistic ~ class, sig, stats::median)
sig$class <- factor(sig$class, levels = medians$class[order(medians$d_statistic)])
medians$class <- factor(medians$class, levels = levels(sig$class))
counts <- table(sig$class)
class_labels <- setNames(paste0(names(counts), " (", counts, ")"), names(counts))

ggplot(sig, aes(x = d_statistic, y = class)) +
  annotate("rect", xmin = -Inf, xmax = 0, ymin = -Inf, ymax = Inf,
           fill = mint, alpha = 0.16) +
  geom_vline(xintercept = c(0, 1), colour = ink, linewidth = 0.4,
             linetype = c("solid", "22")) +
  geom_point(colour = forest, alpha = 0.55, size = 1.9,
             position = position_jitter(height = 0.2, width = 0, seed = 2)) +
  geom_segment(data = medians,
               aes(x = d_statistic, xend = d_statistic,
                   y = as.numeric(class) - 0.34, yend = as.numeric(class) + 0.34),
               colour = coral, linewidth = 1.3) +
  annotate("text", x = 0, y = Inf, label = "inherited by descent", hjust = 1.06,
           vjust = 1.4, size = 3.4, colour = muted) +
  annotate("text", x = 1, y = Inf, label = "random", hjust = 1.15, vjust = 1.4,
           size = 3.4, colour = muted) +
  scale_y_discrete(labels = class_labels, expand = expansion(add = c(0.6, 0.9))) +
  labs(x = "Fritz and Purvis's D", y = NULL) +
  theme_case() +
  theme(panel.grid.major.y = element_blank(),
        axis.text.y = element_text(colour = ink))
```

```{r signal-numbers}
class_median <- function(class) sprintf("%.2f", medians$d_statistic[medians$class == class])
conserved <- sum(sig$distribution == "as conserved as Brownian motion")
```

Most GIFTs are conserved: `r conserved` of the `r nrow(sig)` tested are as
clumped on the tree as descent with modification predicts. The classes differ
in the direction the previous section suggests. Anabolic GIFTs are the most
conserved of the large classes (median D = `r class_median("metabolic, anabolic")`),
catabolic GIFTs are more scattered (`r class_median("metabolic, catabolic")`),
and defense GIFTs are the most scattered of all (`r class_median("defense")`),
as expected of systems that move between lineages.

## Origin adds little beyond lineage

```{r origin, fig.height = 5.6, fig.cap = "Difference associated with each origin group, from phylogenetic regressions in which the twelve groups with at least twenty genomes are fitted together. Points with 95% intervals; filled points pass a false discovery rate of 0.05."}
ori <- origin_traits
ori$origin <- gsub("_", " ", sub("^origin_", "", ori$origin))
response_labels <- c(
  "log2 genome size" = "Genome size (log2)",
  "supported GIFTs" = "Supported GIFTs",
  "deviation from the size expectation (GIFTs)" = "Deviation from expectation"
)
ori$response <- factor(response_labels[ori$response], levels = response_labels)
ori$significant <- ori$q_value < 0.05
ori$origin <- factor(ori$origin, levels = rev(unique(ori$origin)))

ggplot(ori, aes(x = phylogenetic_estimate, y = origin)) +
  geom_vline(xintercept = 0, colour = ink, linewidth = 0.4) +
  geom_segment(aes(x = phylogenetic_estimate - 1.96 * standard_error,
                   xend = phylogenetic_estimate + 1.96 * standard_error,
                   yend = origin), colour = muted, linewidth = 0.6) +
  geom_point(aes(fill = significant), shape = 21, colour = forest, size = 2.8,
             stroke = 0.7) +
  facet_wrap(vars(response), scales = "free_x") +
  scale_fill_manual(values = c(`TRUE` = coral, `FALSE` = surface), guide = "none") +
  labs(x = "difference from genomes without the tag", y = NULL) +
  theme_case() +
  theme(panel.grid.major.y = element_blank(),
        axis.text.y = element_text(colour = ink),
        panel.spacing.x = grid::unit(7, "mm"))
```

```{r origin-numbers}
ori_value <- function(response, origin, column) {
  origin_traits[origin_traits$response == response &
                  origin_traits$origin == origin, column]
}
```

Before accounting for phylogeny, origin groups look very different: plant-associated
genomes support `r round(ori_value("supported GIFTs", "origin_plant_associated", "naive_difference"))`
more GIFTs than the rest and animal-associated genomes
`r abs(round(ori_value("supported GIFTs", "origin_animal_associated", "naive_difference")))`
fewer. Most of that is which lineages live where. In the phylogenetic
regressions two differences remain: plant-associated genomes are larger
(`r round(100 * (2^ori_value("log2 genome size", "origin_plant_associated", "phylogenetic_estimate") - 1))`%),
and genomes from thermal environments support
`r round(ori_value("deviation from the size expectation (GIFTs)", "origin_thermal_geothermal", "phylogenetic_estimate"), 1)`
more GIFTs than their size predicts.

## One step short

A **near miss** is an unsupported GIFT whose closest implementation lacks
exactly one of at least two requirements. gifter never counts a near miss as
support, but it reports each one with the requirement that is missing, and
across a broad panel those reports add up to a map of where the catalogue, the
markers or the biology fall short.

```{r near, fig.height = 6.6, fig.cap = "The fifteen requirements most often missing alone. Each bar is one implementation of a GIFT and the single requirement its near misses lack; the number of genomes that support the GIFT is given at the end of the bar."}
near <- merge(near_steps, gifts[c("gift_id", "gift_name")])
near <- near[order(-near$genomes_one_step_short), ][1:15, ]
near$label <- paste0(gsub("_", " ", near$gift_id), "\n", near$missing_requirement)
near$label <- reorder(near$label, near$genomes_one_step_short)
reading_colours <- c(
  "same step missing, GIFT-specific steps present" = coral,
  "missing step varies, GIFT-specific steps present" = blue,
  "present steps all shared with other GIFTs" = "#b9b8b0"
)

ggplot(near, aes(x = genomes_one_step_short, y = label, fill = reading)) +
  geom_col(width = 0.68) +
  geom_text(aes(label = paste(genomes_supporting_gift, "support")),
            hjust = -0.12, size = 3.2, colour = muted) +
  scale_fill_manual(values = reading_colours, name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.16))) +
  labs(x = "genomes one step short", y = NULL) +
  guides(fill = guide_legend(ncol = 1)) +
  theme_case() +
  theme(panel.grid.major.y = element_blank(),
        axis.text.y = element_text(colour = ink, size = 9, lineheight = 0.9),
        legend.justification = "left")
```

```{r near-numbers}
reading_share <- function(reading) {
  round(100 * near_readings$share_of_near_miss_calls[near_readings$reading == reading])
}
top <- near_steps[which.max(near_steps$genomes_one_step_short), ]
```

The median genome is one step short of `r median(near_genomes$near_misses)`
GIFTs. The near misses fall into three readings:

| Reading | Share of near misses | What it suggests |
|---|---|---|
| The same step is missing, and steps specific to the GIFT are present | `r reading_share("same step missing, GIFT-specific steps present")`% | An alternative enzyme or marker we have not curated yet, or a step the lineage does not encode |
| The missing step varies, and steps specific to the GIFT are present | `r reading_share("missing step varies, GIFT-specific steps present")`% | Scattered gaps, with no single candidate |
| Every step present is shared with other GIFTs | `r reading_share("present steps all shared with other GIFTs")`% | No evidence for this GIFT in particular; the shared steps belong to something else |

The first reading is the one we can act on. The most frequent case is
`r gsub("_", " ", top$gift_id)`: `r top$genomes_one_step_short` genomes carry
one of its two requirements and lack `r top$missing_requirement`, while only
`r top$genomes_supporting_gift` support the GIFT. A gap that consistent is a point we
will analyse to improve the definition of the GIFT.

We have logged these gaps as open leads, and none has been assessed yet. Each
one will go through the full curation process, which may add an alternative
route or marker, or confirm that the lineage does not encode the step. The
panel is then evaluated again with the new database release, so the numbers on
this page belong to database `r item(audit, "database version")` and will move
as the definitions improve. The loop is described in
[Refine the catalogue iteratively](curating-a-gift.html#refine-the-catalogue-iteratively).

## Reproduce this analysis

The panel, the calls and every table behind this page are committed to the
repository under
[`manuscript/analysis/`](https://github.com/alberdilab/gifter/tree/main/manuscript/analysis).

| Script | What it does |
|---|---|
| `23-gtdb-phylogeny.R` | Selects and locks the panel, evaluates the calls |
| `26-gtdb-repertoire-genome-size.R` | Size expectation and deviations by taxon |
| `27-gtdb-size-signal-and-gift-classes.R` | Phylogenetic signal of the deviation; classes of GIFT against size |
| `28-gtdb-origin-and-annotations.R` | Origin groups and assembly annotations |
| `29-gtdb-near-misses-and-gift-signal.R` | Near misses; phylogenetic signal of each GIFT |
| `30-gtdb-near-miss-steps.R` | Which requirement is missing in a near miss |

The panel itself, with accessions, taxonomy, origin tags and the pruned tree,
is in
[`gtdb-phylogeny/`](https://github.com/alberdilab/gifter/tree/main/manuscript/analysis/gtdb-phylogeny).
